% REFERENCE.TEX
%
% The documentation in this file is part of PyXPlot
% <http://www.pyxplot.org.uk>
%
% Copyright (C) 2006-2010 Dominic Ford <coders@pyxplot.org.uk>
%               2009-2010 Ross Church
%
% $Id$
%
% PyXPlot is free software; you can redistribute it and/or modify it under the
% terms of the GNU General Public License as published by the Free Software
% Foundation; either version 2 of the License, or (at your option) any later
% version.
%
% You should have received a copy of the GNU General Public License along with
% PyXPlot; if not, write to the Free Software Foundation, Inc., 51 Franklin
% Street, Fifth Floor, Boston, MA  02110-1301, USA

% ----------------------------------------------------------------------------

% LaTeX source for the PyXPlot Users' Guide

\chapter{Command Reference}
\label{ch:reference}

This chapter contains an alphabetically ordered list of all of PyXPlot's
commands. The syntax of each is specified in a variant of Backus-Naur notation,
in which angle brackets {\tt <>} are used to indicate replaceable tokens, curly
brackets {\tt ()} are used to indicate mutually-exclusive options which are
separated by vertical lines {\tt |}, square brackets {\tt []} are used to
indicate optional items, and braces {\tt \{\}} are used to indicate items which
may be repeated. Dots {\tt ...} are used to indicate arbitrary strings of text.
Where any of these punctuation marks appear as a part of PyXPlot's syntax, they
are placed in double quotes, as in {\tt "\{"}. Strings, such as filenames,
which should be placed in quotes are placed in single quotes, as in {\tt
'<filename>'}, though in practice, double quotes can always be used
interchangeably with single quotes. In cases where there is likely to be any
ambiguity, worked examples are usually shown beneath the syntax specification.

\section{?}\indcmd{?}

\begin{verbatim}
? [<topic> {<sub-topic>} ]
\end{verbatim}

The {\tt ?} symbol is a shortcut to the {\tt help} command.


\section{!}\indcmd{!}

\begin{verbatim}
! <shell command>
... `<shell command>` ...
\end{verbatim}

Shell commands can be executed within PyXPlot by prefixing them with
pling (!) characters, as in the example:

\begin{verbatim}
!mkdir foo
\end{verbatim}

\noindent As an alternative, back-quotes (`) can be used to substitute the
output of a shell command into a PyXPlot command, as in the example:

\begin{verbatim}
set xlabel `echo "'" ; ls ; echo "'"`
\end{verbatim}

\noindent Note that back-quotes cannot be used inside quote characters, and so
the following would \textit{not} work:

\begin{verbatim}
set xlabel '`ls`'
\end{verbatim}


\section{arc}\indcmd{arc}

\begin{verbatim}
arc [ item <id> ] [at] <x>, <y> radius <r>
    from <start> to <finish> [ with {<option>} ]
\end{verbatim}

Arcs (curves with constant radius of curvature, that is, segments of circles)
may be drawn on multiplot canvases using the \indcmdt{arc}.  The {\tt at}
modifier specifies the coordinates of the centre of curvature, from which all
points on the arc are at the distance given following the {\tt radius} modifier.
The angles {\tt start} and {\tt finish}, measured clockwise from the vertical,
control where the arc begins and finishes.  For example, the command

\begin{verbatim}
arc at 0,0 radius 2 from 90 to 270
\end{verbatim}

\noindent would draw a semi-circle beneath the line $x=0$, centred on the
origin with radius $2\,\mathrm{cm}$.  The usual style modifiers for lines may
be passed after the keyword {\tt with}; if the {\tt fillcolour} modifier is
specified then the arc will be filled to form a pie-chart slice.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{arrow}\indcmd{arrow}

\begin{verbatim}
arrow [ item <id> ] [from] <x>, <y> [to] <x>, <y>
                  [ with {<option>} ]
\end{verbatim}

Arrows may be drawn on multiplot canvases using the \indcmdt{arrow}. The style
of the arrows produced may be specified by following the \indmodt{with}
modifier by one of the style keywords \indkeyt{nohead}, \indkeyt{head}
(default) or \indkeyt{twohead}. In addition, keywords such as \indkeyt{colour},
\indkeyt{linewidth} and \indkeyt{linetype} have the same syntax and meaning
following the \indmodt{with} keyword as in the {\tt plot} command. The
following example would draw a bidirectional blue arrow:

\begin{verbatim}
arrow from x1,y1 to x2,y2 with twohead linetype 2 colour blue
\end{verbatim}

The {\tt arrow} command has a twin, the {\tt line} command, which has the same
syntax, but uses the default arrow style of \indkeyt{nohead}, producing short
line segments.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{box}\indcmd{box}

\begin{verbatim}
box [ item <id> ] at <x>, <y> width <w> height <h>
         [ rotate <r> ] [ with {<option>} ]

box [ item <id> ] from <x1>, <y1> to <x2>, <y2>
         [ rotate <r> ] [ with {<option>} ]
\end{verbatim}

The \indcmdt{box} may be used to draw and fill rectangular boxes on multiplot
canvases.  The position of each box may be specified in one of two ways.  In the
first, the coordinates of one corner of the box are specified, along with its
width and height. If both the width and the height are positive then the
coordinates are taken to be those of the bottom left-hand corner of the box;
other corners may be specified if the supplied width and/or height are
negative. If a rotation angle is specified then the box is rotated about the
specified corner.  The {\tt with} modifier allows the style of the box to be
specified using similar options to those accepted by the {\tt plot} command.

The second syntax allows two pairs of coordinates to be specified.  PyXPlot
will then draw a rectangular box with opposing corners at the specified
locations.  If an angle is specified the box will be rotated about its centre.
Hence the following two commands both draw a square box centred on the origin:

\begin{verbatim}
box from -1, -1 to 1,1
box at 1, -1 width -2 height 2
\end{verbatim}

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{break}\indcmd{break}

\begin{verbatim}
break [<loopname>]
\end{verbatim}

The \indcmdt{break} terminates execution of {\tt do}, {\tt while}, {\tt for}
and {\tt foreach} loops in an analogous manner to the {\tt break} statement in
the C programming language.  Execution resumes at the statement following the
end of the loop. For example, the following loop would only print the numbers~1
and~2:

\begin{verbatim}
for i = 1 to 10
 {
  print i
  if (i==2)
   {
    break
   }
 }
\end{verbatim}

If several loops are nested, the {\tt break} statement only acts on the
innermost loop. If the {\tt break} statement is encountered outside of any loop
structure, an error results. Optionally, the {\tt for}, {\tt foreach}, {\tt do}
and {\tt while} commands may be supplied with a name for the loop, prefixed by
the word {\tt loopname}, as in the examples:

\begin{verbatim}
for i=0 to 4 loopname iloop

foreach i in "*.dat" loopname DatafileLoop
\end{verbatim}

\noindent When loops are given such names, the {\tt break} statement may be
followed by the name of the loop whose iteration is to be broken, allowing it
to act upon loops other than the innermost one.

See also the {\tt continue} command.


\section{cd}\indcmd{cd}

\begin{verbatim}
cd <directory>
\end{verbatim}

PyXPlot's \indcmdt{cd} is very similar to the shell {\tt cd} command; it
changes the current working directory. The following example would enter the
subdirectory {\tt foo}:

\begin{verbatim}
cd foo
\end{verbatim}


\section{circle}\indcmd{circle}

\begin{verbatim}
circle [ item <id> ] [at] <x>, <y> radius <r>
       [ with {<option>} ]
\end{verbatim}

The \indcmdt{circle} is used to draw circles on multiplot canvases.  The
coordinates of the circle's centre and its radius are specified. The {\tt with}
modifier allows the style of the circle to be specified using similar options
to those accepted by the {\tt plot} command.  The example

\begin{verbatim}
circle at 2,2 radius 1 with colour red fillcolour blue
\end{verbatim}

\noindent would draw a red circle of unit radius filled in blue, centred
$2\,\mathrm{cm}$ above and to the right of the origin.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{clear}\indcmd{clear}

\begin{verbatim}
clear
\end{verbatim}

In multiplot mode, the \indcmdt{clear} removes all plots, arrows and text
objects from the working multiplot canvas. Outside of multiplot mode, it is not
especially useful; it removes the current plot to leave a blank canvas.  The
{\tt clear} command should not be followed by any parameters.


\section{continue}\indcmd{continue}

\begin{verbatim}
continue [<loopname>]
\end{verbatim}

The \indcmdt{continue} terminates execution of the current iteration of {\tt
for}, {\tt foreach}, {\tt do} and {\tt while} loops in an analogous manner to
the {\tt continue} statement in the C programming language. Execution resumes
at the first statement of the next iteration of the loop, or at the first
statement following the end of the loop in the case of the last iteration of
the loop.  For example, the following script will not print the number~2:

\begin{verbatim}
for i = 0 to 5
 {
  if (i==2)
   {
    continue
   }
  print i
 }
\end{verbatim}

If several loops are nested, the {\tt continue} statement only
acts on the innermost loop. If the {\tt continue} statement is encountered outside of any
loop structure, an error results. Optionally, the {\tt for}, {\tt foreach},
{\tt do} and {\tt while} statements may be supplied with a name for the loop, prefixed by
the word {\tt loopname}, as in the examples:

\begin{verbatim}
for i=0 to 4 loopname iloop

foreach i in "*.dat" loopname DatafileLoop
\end{verbatim}

\noindent When loops are given such names, the {\tt continue} statement may be
followed by the name of the loop whose iteration is to be broken, allowing it
to act upon loops other than the innermost one.

See also the {\tt break} command.


\section{delete}\indcmd{delete}

\begin{verbatim}
delete <item number> {, <item number>}
\end{verbatim}

The \indcmdt{delete} removes vector graphics objects such as plots, arrows or
text items from the current multiplot canvas. All vector graphics objects
placed on multiplot canvases receive unique identification numbers which count
sequentially from one, and which may be listed using the {\tt list} command.
The items to be deleted should be identified using a comma-separated list of
their identification numbers. The example

\begin{verbatim}
delete 1,2,3
\end{verbatim}

\noindent would remove item numbers~1,~2 and~3.

Having been deleted, multiplot items can be restored using the {\tt undelete}
command.


\section{do}\indcmd{do}

\begin{verbatim}
do [loopname <loopname>]
 "{"
  ...
 "}" while <condition>
\end{verbatim}

The \indcmdt{do} executes a block of commands repeatedly, checking the
condition given in the {\tt while} clause at the end of each iteration.  If the
condition is true then the loop executes again. This is similar to a {\tt
while} loop, except that the contents of a {\tt do} loop are {\emph always}
executed at least once.  The following example prints the numbers~1, 2 and~3:

\begin{verbatim}
i=1
do
 {
  print i
  i = i + 1
 } while (i < 4)
\end{verbatim}

\noindent Note that there must always be a newline following the opening brace
after the \indcmdt{do}, and the while clause must always be on the same line as
the closing brace.


\section{ellipse}\indcmd{ellipse}

Ellipses may be drawn on multiplot canvases using the \indcmdt{ellipse}. The shape
of the ellipse may be specified in many different ways, by specifying

\begin{enumerate}[(i)]
\item the position of two corners of the smallest rectangle which can enclose
the ellipse when its major axis is horizontal, together with an optional
counter-clockwise rotation angle, applied about the centre of the ellipse.
For example:

\begin{verbatim}
ellipse from 0,0 to 4,1 rot 70
\end{verbatim}

\item the position of both the centre and one of the foci of the ellipse,
together with any one of the following additional pieces of information: the
ellipse's major axis length, its semi-major axis length, its minor axis length,
its semi-minor axis length, its eccentricity, its latus rectum, or its
semi-latus rectum.  For example:

\begin{verbatim}
ellipse focus 0,0 centre 2,2 majoraxis 4
ellipse focus 0,0 centre 2,2 minoraxis 4
ellipse focus 0,0 centre 2,2 ecc 0.5
ellipse focus 0,0 centre 2,2 LatusRectum 6
ellipse focus 0,0 centre 2,2 slr 3
\end{verbatim}

\item the position of either the centre or one of the foci of the ellipse,
together with any two of the following additional pieces of information: the
ellipse's major axis length, its semi-major axis length, its minor axis length,
its semi-minor axis length, its eccentricity, its latus rectum, or its
semi-latus rectum. An optional counter-clockwise rotation angle may also be
specified, applied about either the centre or one of the foci of the ellipse,
whichever is specified. If no rotation angle is given, then the major axis of
the ellipse is horizontal.  For example:

\begin{verbatim}
ellipse centre 0,0 majoraxis 4 minoraxis 4
\end{verbatim}
\end{enumerate}

The line type, line width, and colour of line with which the outlines of
ellipses are drawn may be specified after the keyword {\tt with}, as in the
{\tt box} and {\tt circle} commands above. Likewise, ellipses may be filled in
the same manner.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{else}\indcmd{else}

The {\tt else} statement is described in the entry for the {\tt if}
statement, of which it forms part.


\section{eps}\indcmd{eps}

\begin{verbatim}
eps [ item <id> ] '<filename>' [at <x>, <y>] [rotate <angle>]
                            [width <width>] [height <height>]
\end{verbatim}

The \indcmdt{eps} allows encapsulated postscript (eps) images to be inserted
onto multiplot canvases.  The {\tt at} modifier can be used to specify where
the bottom-left corner of the image should be placed; if it is not, then the
image is placed at the origin. The {\tt rotate} modifier can be used to rotate
the image by any angle, measured in degrees counter-clockwise.  The {\tt width}
or {\tt height} modifiers can be used to specify the width or height with which
the image should be rendered; both should be specified in centimetres. If
neither is specified then the image will be rendered with the native dimensions
specified within the postscript.  The {\tt eps} command is often useful in
multiplot mode, allowing postscript images to be combined with plots, text
labels, etc.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{exec}\indcmd{exec}

\begin{verbatim}
exec <command>
\end{verbatim}

The \indcmdt{exec} can be used to execute PyXPlot commands contained within
string variables, as in the following example:

\begin{verbatim}
terminal="eps"
exec "set terminal %s"%(terminal)
\end{verbatim}

\noindent It can also be used to write obfuscated PyXPlot scripts, and its use
should be minimised wherever possible.


\section{exit}\indcmd{exit}

\begin{verbatim}
exit
\end{verbatim}

The \indcmdt{exit} can be used to quit PyXPlot. If multiple command files,
or a mixture of command files and interactive sessions, were specified on
PyXPlot's command line, then PyXPlot moves onto the next command-line item
after receiving the {\tt exit} command.

PyXPlot may also be quit be pressing CTRL-D or via the {\tt quit} command. In
interactive mode, CTRL-C terminates the current command, if one is running.
When running a script, CTRL-C terminates execution of the script.


\section{fft}\indcmd{fft}

\begin{verbatim}
fft {<range>} <function>"()"
    of ( '<filename>' | <function>"()" )
    [using <expression> {:<expression>} ]

ifft {<range>} <function>"()"
    of ( '<filename>' | <function>"()" )
    [using <expression> {:<expression>} ]
\end{verbatim}

The \indcmdt{fft} calculates Fourier transforms of \datafile s or functions.
Transforms can be performed on datasets with arbitrary numbers of dimensions.
To transform an algebraic expression with $n$~degrees of freedom, it must be
wrapped in a function of the form $f(i_2,i_2,\ldots,i_n)$. To transform an
$n$-dimensional dataset stored in a \datafile, the samples must be arranged on
a regular linearly-spaced grid and stored in row-major order.  For each
dimension of the transform, a range specification must be provided to the {\tt
fft} command in the form

\begin{verbatim}
[ <minimum> : <maximum> : <step> ]
\end{verbatim}

When data from a \datafile\ is being transformed, the specified range(s) must
precisely match those of the samples read from the file; the first $n$~columns
of data should contain the values of the $n$~real-space coordinates, and the
$n+1$th column should contain the data to be transformed.  After the range(s),
a function name should be provided for the output transform: a function of
$n$~arguments with this name will be generated to represent the transformed
data.  Note that this function is in general complex -- i.e.\ it has a non-zero
imaginary component. Complex numerics can be enabled using the {\tt set
numerics complex} command and the {\tt fft} command is of little use without
doing so. The {\tt using}, {\tt index}, {\tt every} and {\tt select} modifiers
can be used to specify how data will be sampled from the input function or
\datafile\ in an analogous manner to how they are used in the {\tt plot}
command.

The {\tt ifft} command calculates inverse Fourier transforms; it has the same
syntax as the {\tt fft} command.

\section{fit}\indcmd{fit}

\begin{verbatim}
fit [{<range>}] <function>"()" [withouterrors] '<datafile>'
    [index <value>]
    [using <expression> {:<expression>} ]
    via <variable> {, <variable>}
\end{verbatim}

The \indcmdt{fit} can be used to fit arbitrary functional forms to \datapoint s
read from files. It can be used to produce best-fit lines for datasets or to
determine gradients and other mathematical properties of data by looking at the
parameters associated with the best-fitting functional form.  The following
simple example fits a straight line to data in a file called {\tt data.dat}:

\begin{verbatim}
f(x) = a*x+b
fit f() 'data.dat' index 1 using 2:3 via a,b
\end{verbatim}

\noindent The first line specifies the functional form which is to be used.
The coefficients within this function which are to be varied during the fitting
process are listed after the keyword \indkeyt{via} in the {\tt fit} command.
The modifiers \indmodt{index}, \indmodt{every}, \indmodt{select} and
\indmodt{using} have the same meanings as in the {\tt plot} command.

As the {\tt fit} command works, it displays statistics including the best-fit
values of each of the fitting parameters, the uncertainties in each of them,
and the covariance matrix. These can be useful for analysing the security of
the fit achieved, but calculating the uncertainties in the best-fit parameters
and the covariance matrix can be time consuming, especially when many
parameters are being fitted simultaneously. The optional keyword {\tt
withouterrors} can be included immediately before the filename of the
\datafile\ to be fitted to substantially speed up cases where this information
is not required.

More details can be found in Section~\ref{sec:fit_command}.


\section{for}\indcmd{for}

\begin{verbatim}
for <variable> = <start> to <end> [step <step>]
                          [loopname <loopname>]
 "{"
  ...
 "}"
\end{verbatim}

The \indcmdt{for} executes a set of commands repeatedly, with a specified
variable taking a different value on each iteration. The variable takes the
value {\tt start} on the first iteration, and increases by a fixed value {\tt
step} on each iteration; {\tt step} may be negative if {\tt end} $<$ {\tt
start}. If {\tt step} is not specified then a value of unity is assumed. The
loop terminates when the variable exceeds or equals {\tt end}; note that the
loop is not executed with the variable equalling {\tt end}.  The following
example prints the squares of the first five natural numbers:

\begin{verbatim}
for i = 1 to 6
 {
  print i**2
 }
\end{verbatim}


\section{foreach}\indcmd{foreach}

\begin{verbatim}
foreach <variable> in ( <filename expression> |
                         "("<value> {, <value>}")" )
                      [loopname <loopname>]
 "{"
  ...
 "}"
\end{verbatim}

The \indcmdt{foreach} can be used to run a block of commands repeatedly, once
for each item in a list.  The list of items can be specified in one of two
ways.  In the first case, a set of filenames or filename wildcards is supplied,
and the {\tt foreach} loop iterates once for each supplied filename, with a
string variable set to each filename in succession.  For example, the following
loop would plot the data in the set of files whose names end with {\tt .dat}:

\begin{verbatim}
plot     # Create blank plot
foreach file in *.dat
 {
  replot file with lines
 }
\end{verbatim}

The second form of the command takes a list of string or numerical values
provided explicitly by the user, and the {\tt foreach} loop iterates once for
each value, with a variable set to each value in succession.  For example, the
following script would plot normal distributions of three different widths:

\begin{verbatim}
plot     # Create blank plot
foreach sigma in (1, 2, 3)
 {
  replot 1/sigma*exp(-x**2/(2*sigma**2))
 }
\end{verbatim}


\section{help}\indcmd{help}

\begin{verbatim}
help [<topic> {<sub-topic>} ]
\end{verbatim}

The \indcmdt{help} provides an hierarchical source of information which is
supplementary to this Users' Guide.  To obtain information on any particular
topic, type {\tt help} followed by the name of the topic, as in the following
example

\begin{verbatim}
help commands
\end{verbatim}

\noindent which provides information on PyXPlot's commands. Some topics have
sub-topics; these are listed at the end of each help page. To view them, add
further words to the end of the help request, as in the example:

\begin{verbatim}
help commands help
\end{verbatim}

Information is arranged with general information about PyXPlot under the
heading {\tt about} and information about PyXPlot's commands under {\tt
commands}.  Information about the format that input \datafile s should take can
be found under {\tt datafile}.  Other categories are self-explanatory.

To exit any help page, press the {\tt q} key.


\section{histogram}\indcmd{histogram}

\begin{verbatim}
histogram [<range>] <function name>"()" '<datafile>'
     [every <expression> {:<expression>} ]
     [index <value>]
     [select <expression>]
     [using <expression> {:<expression>} ]
     ( [binwidth <value>] [binorigin <value>] |
       [bins (x1, x2, ...)] )
\end{verbatim}

The \indcmdt{histogram} takes a single column of data from a file and produces
a function that represents the frequency distribution of the supplied data
values. The output function consists of a series of discrete intervals which we
term {\it bins}. Within each interval the output function has a constant value,
determined such that the area under each interval -- i.e.\ the integral of the
function over each interval -- is equal to the number of datapoints found
within that interval.  The following simple example

\begin{verbatim}
histogram f() 'input.dat'
\end{verbatim}

\noindent produces a frequency distribution of the data values found in the
first column of the file {\tt input.dat}, which it stores in the function
$f(x)$. The value of this function at any given point is equal to the number of
items in the bin at that point, divided by the width of the bins used. If the
input datapoints are not dimensionless then the output frequency distribution
adopts appropriate units, thus a histogram of data with units of length has
units of one over length.

The number and arrangement of bins used by the \indcmdt{histogram} can be
controlled by means of various modifiers.  The \indmodt{binwidth} modifier sets
the width of the bins used. The \indmodt{binorigin} modifier controls where
their boundaries lie; the \indcmdt{histogram} selects a system of bins which,
if extended to infinity in both directions, would put a bin boundary at the
value specified in the {\tt binorigin} modifier. Thus, if {\tt binorigin 0.1}
were specified, together with a bin width of~20, bin boundaries might lie
at~$20.1$, $40.1$, $60.1$, and so on. Alternatively global defaults for the bin
width and the bin origin can be specified using the {\tt set binwidth} and {\tt
set binorigin} commands respectively. The example

\begin{verbatim}
histogram h() 'input.dat' binorigin 0.5 binwidth 2
\end{verbatim}

\noindent would bin data into bins between $0.5$ and $2.5$, between $2.5$ and
$4.5$, and so forth.

Alternatively the set of bins to be used can be controlled more precisely using
the \indmodt{bins} modifier, which allows an arbitrary set of bins to be
specified. The example

\begin{verbatim}
histogram g() 'input.dat' bins (1, 2, 4)
\end{verbatim}

\noindent would bin the data into two bins, $x=1\to 2$ and $x=2\to 4$.

A range can be supplied immediately following the {\tt histogram} command,
using the same syntax as in the {\tt plot} and {\tt fit} commands; if such a
range is supplied, only points that fall within that range will be binned.  In
the same way as in the {\tt plot} command, the \indmodt{index},
\indmodt{every}, \indmodt{using} and \indmodt{select} modifiers can be used to
specify which subsets of a \datafile\ should be used.

Two points about the {\tt histogram} command are worthy of note. First,
although histograms are similar to bar charts, they are not the same.  A bar
chart conventionally has the height of each bar equal to the number of points
that it represents, whereas a histogram is a continuous function in which the
area underneath each interval is equal to the number of points within it.
Thus, to produce a bar chart using the {\tt histogram} command, the end result
should be multiplied by the bin width used.

Second, if the function produced by the {\tt histogram} command is plotted
using the {\tt plot} command, samples are automatically taken not at evenly
spaced intervals along the ordinate axis, but at the centres of each bin. If
the \indpst{boxes} plot style is used, the box boundaries are be conveniently
drawn to coincide with the bins into which the data were sorted.


\section{history}\indcmd{history}

\begin{verbatim}
history [<number of items>]
\end{verbatim}

The \indcmdt{history} prints a list of the most recently executed commands on
the terminal.  The optional parameter, {\tt N}, if supplied, causes only the
latest $N$ commands to be displayed.


\section{if}\indcmd{if}\indcmd{else if}\indcmd{else}

\begin{verbatim}
if <criterion>
 "{"
    ...
 "}" { else if <criterion>
 "{"
    ...
 "}" } [ else
 "{"
    ...
 "}" ]
\end{verbatim}

The \indcmdt{if} allows conditional execution of blocks of commands.  The code
enclosed in braces following the {\tt if} statement is executed if, and only
if, the {\tt criterion} is satisfied.  An arbitrary number of subsequent {\tt
else if} statements can optionally follow the initial {\tt if} statement; these
have their own criteria for execution which are only considered if all of the
previous criteria have tested false -- i.e.\ if none of the previous command
blocks have been executed.  A final optional {\tt else} statement can be
provided; the block of commands which follows it are executed only if none of
the preceding criteria have tested true.  The following example illustrates a
chain of {\tt else if} clauses:

\begin{verbatim}
if (x==2)
 {
  print "x is two!"
 } else if (x==3) {
  print "x is three!"
 } else if (x>3) {
  print "x is greater than three!"
 } else {
  x=2
  print "x didn't used to be two, but it is now!"
 }
\end{verbatim}


\section{ifft}\indcmd{ifft}

\begin{verbatim}
ifft {<range>} <function>"()"
    of ( '<filename>' | <function>"()" )
    [using <expression> {:<expression>} ]
\end{verbatim}

See {\tt fft}.


\section{image}\indcmd{image}

\begin{verbatim}
image [ item <id> ] '<filename>' [at <x>, <y>] [rotate <angle>]
                     [width <width>] [height <height>] [smooth]
                   [notransparent] [transparent rgb<r>:<g>:<b>]
\end{verbatim}

The \indcmdt{image} allows graphical images to be inserted onto the current
multiplot canvas from files on disk. Input graphical images may be in bitmap,
gif, jpeg or png formats; the file type of each image is automatically
detected. The {\tt at} modifier can be used to specify where the bottom-left
corner of each image should be placed; if no position is specified then this
corner of the image is placed at the origin. The {\tt rotate} modifier can be
used to rotate images by any angle, measured in degrees counter-clockwise.  The
{\tt width} or {\tt height} modifiers can be used to specify the width or
height with which images should be rendered; both should be specified in
centimetres. If neither is specified then images are rendered with the native
dimensions specified within the metadata present in the image file (if any). If
both are specified, then the aspect ratio of the image is changed.

The keyword {\tt smooth} may optionally be supplied to cause the pixels of
images to be interpolated\footnote{Many commonly-used postscript display
engines, including GhostScript, do not support this functionality.}.  Images
which include transparency are supported. The optional keyword {\tt
notransparent} may be supplied to cause transparent regions to be filled with
the image's default background colour. Alternatively, an RGB colour may be
specified in the form {\tt rgb<r>:<g>:<b>} after the keyword {\tt transparent}
to cause that particular colour to become transparent; the three components of
the RGB colour should be in the range~0 to~255.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{interpolate}\indcmd{interpolate}

\begin{verbatim}
interpolate ( akima | linear | loglinear | polynomial | spline )
            [<range specification>] <function name>"()"
            '<filename>'
            [ every <expression> {:<expression} ]
            [ index <value> ]
            [ select <expression> ]
            [ using <expression> {:<expression} ]
\end{verbatim}

The \indcmdt{interpolate} can be used to generate a special function within
PyXPlot's mathematical environment which interpolates a set of \datapoint s
supplied from a \datafile.  Various different types of interpolation are
supported: linear interpolation, power-law interpolation, polynomial
interpolation, cubic-spline interpolation and akima-spline interpolation. The
use of polynomial interpolation with large datasets is strongly discouraged, as
polynomial fits tend to show severe oscillations between \datapoint s.
Extrapolation is not permitted; if an attempt is made to evaluate an
interpolated function beyond the limits of the \datapoint s which it
interpolates, PyXPlot returns an error or value of not-a-number.

A very common application of the \indcmdt{interpolate} is to perform arithmetic
functions such as addition or subtraction on datasets which are not sampled at
the same abscissa values. The following example would plot the difference
between two such datasets:

\begin{verbatim}
interpolate linear f() 'data1.dat'
interpolate linear g() 'data2.dat'
plot [min:max] f(x)-g(x)
\end{verbatim}

\noindent Note that it is advisable to supply a range to the {\tt plot} command
in this example: because the two datasets have been turned into continuous
functions, the {\tt plot} command has to guess a range over which to plot them
unless one is explicitly supplied.

The \indcmdt{spline} is an alias for {\tt interpolate spline}; the following
two statements are equivalent:

\begin{verbatim}
spline f() 'data1.dat'
interpolate spline f() 'data1.dat'
\end{verbatim}


\section{jpeg}\indcmd{jpeg}

\begin{verbatim}
jpeg [ item <id> ] '<filename>' [at <x>, <y>] [rotate <angle>]
                    [width <width>] [height <height>] [smooth]
                  [notransparent] [transparent rgb<r>:<g>:<b>]
\end{verbatim}

See {\tt image}.


\section{let}\indcmd{let}

\begin{verbatim}
let <varname> = <value>
\end{verbatim}

The \indcmdt{let} sets the variable {\tt varname} to equal {\tt value}.


\section{list}\indcmd{list}

\begin{verbatim}
list
\end{verbatim}

The \indcmdt{list} returns a list of all of the items on the current multiplot
canvas, giving their identification numbers and the commands used to produce
them.  The following is an example of the output produced:

{\footnotesize
\begin{verbatim}
pyxplot> list
# ID   Command
    1  plot item 1 f(x) using columns
    2  [deleted] text item 2 "Figure 1: A plot of f(x)" at 0,0 rotate 0 gap 0
    3  text item 3 "Figure 1: A plot of $f(x)$" at 0,0 rotate 0 gap 0
\end{verbatim}
}

In this example, the user has plotted a graph of $f(x)$ and added a caption to
it. The {\tt ID} column lists the reference numbers of each multiplot item.
Item {\tt 1} has been deleted.


\section{load}\indcmd{load}

\begin{verbatim}
load '<filename>'
\end{verbatim}

The \indcmdt{load} executes a PyXPlot command script file, just as if its
contents had been typed into the current terminal. For example:

\begin{verbatim}
load 'foo'
\end{verbatim}

\noindent would have the same effect as typing the contents of the file {\tt
foo} into the present terminal.  Filename wildcard can be supplied, in which
case {\it all} command files matching the given wildcard are executed, as in
the example:

\begin{verbatim}
load '*.script'
\end{verbatim}


\section{maximise}\indcmd{maximise}

\begin{verbatim}
maximise <expression> via <variable> {, variable}
\end{verbatim}

The \indcmdt{maximise} can be used to find the maxima of algebraic expressions.
A single algebraic expression should be supplied for optimisation, together
with a comma-separated list of the variables which it should be optimised with
respect to. In the following example, a maximum of the sinusoidal function
$\cos(x)$ is sought:

\vspace{3mm}
\noindent{\tt pyxplot> {\bf set numerics real}}\newline
\noindent{\tt pyxplot> {\bf x=0.1}}\newline
\noindent{\tt pyxplot> {\bf maximise cos(x) via x}}\newline
\noindent{\tt pyxplot> {\bf print x/pi}}\newline
\noindent{\tt 0}
\vspace{3mm}

\noindent Note that this particular example doesn't work when complex
arithmetic is enabled, since $\cos(x)$ diverges to $\infty$ at $x=\infty
i$, and this is correctly identified by PyXPlot as the global maximum of the
function when complex numbers are allowed.

Various caveats apply the {\tt maximise} command, as well as to the {\tt
minimise} and {\tt solve} commands.  All of these commands operate by searching
numerically for optimal sets of input parameters to meet the criteria set by
the user. As with all numerical algorithms, there is no guarantee that the {\it
locally} optimum solutions returned are the {\it globally} optimum solutions.
It is always advisable to double-check that the answered returned agree with
common sense.

These commands can often find solutions to equations when these solutions are
either very large or very small, but they usually work best when the solution
they are looking for is roughly of order unity.  PyXPlot does have mechanisms
which attempt to correct cases where the supplied initial guess turns out to be
many orders of magnitude different from the true solution, but it cannot be
guaranteed not to wildly overshoot and produce unexpected results in such
cases.  To reiterate, it is always advisable to double-check that the answered
returned agree with common sense.


\section{minimise}\indcmd{minimise}

\begin{verbatim}
minimise <expression> via <variable> {, variable}
\end{verbatim}

The \indcmdt{minimise} can be used to find the minima of algebraic expressions.
A single algebraic expression should be supplied for optimisation, together
with a comma-separated list of the variables which it should be optimised with
respect to. In the following example, a minimum of the sinusoidal function
$\cos(x)$ is sought:

\vspace{3mm}
\noindent{\tt pyxplot> {\bf set numerics real}}\newline
\noindent{\tt pyxplot> {\bf x=0.1}}\newline
\noindent{\tt pyxplot> {\bf minimise cos(x) via x}}\newline
\noindent{\tt pyxplot> {\bf print x/pi}}\newline
\noindent{\tt 1}
\vspace{3mm}

\noindent Note that this particular example doesn't work when complex
arithmetic is enabled, since $\cos(x)$ diverges to $-\infty$ at $x=\pi+\infty
i$, and this is correctly identified by PyXPlot as the global minimum of the
function when complex numbers are allowed.

Various caveats apply the {\tt minimise} command, as well as to the {\tt
maximise} and {\tt solve} commands.  All of these commands operate by searching
numerically for optimal sets of input parameters to meet the criteria set by
the user. As with all numerical algorithms, there is no guarantee that the {\it
locally} optimum solutions returned are the {\it globally} optimum solutions.
It is always advisable to double-check that the answered returned agree with
common sense.

These commands can often find solutions to equations when these solutions are
either very large or very small, but they usually work best when the solution
they are looking for is roughly of order unity.  PyXPlot does have mechanisms
which attempt to correct cases where the supplied initial guess turns out to be
many orders of magnitude different from the true solution, but it cannot be
guaranteed not to wildly overshoot and produce unexpected results in such
cases.  To reiterate, it is always advisable to double-check that the answered
returned agree with common sense.


\section{move}\indcmd{move}

\begin{verbatim}
move <item number> to <x>, <y>
\end{verbatim}

The \indcmdt{move} allows vector graphics objects to be moved around on the
current multiplot canvas. All vector graphics objects placed on multiplot
canvases receive unique identification numbers which count sequentially from
one, and which may be listed using the {\tt list} command. The item to be moved
should be specified using its identification number. The example

\begin{verbatim}
move 23 to 8,8
\end{verbatim}

\noindent would move multiplot item~23 to position $(8,8)$ centimetres. If this
item were a plot, the end result would be the same as if the command {\tt set
origin 8,8} had been executed before it had originally been plotted.


\section{plot}\indcmd{plot}

\begin{verbatim}
plot [item <id>] [{<range>}] ( '<filename>' | <function> )
     [axes <axes>] [every <expression> {:<expression>} ]
     [index <value>] [select <expression>]
     [title <string>] [using <expression> {:<expression>} ]
     [with {<option>} ]
\end{verbatim}

The \indcmdt{plot} is used to produce graphs. The following simple example
would plot the sine function:

\begin{verbatim}
plot sin(x)
\end{verbatim}

Ranges for the axes of a graph can be specified by placing them in
square brackets before the name of the function to be plotted. An example of
this syntax would be:

\begin{verbatim}
plot [-pi:pi] sin(x)
\end{verbatim}

\noindent which would plot the function $\sin(x)$ between $-\pi$ and $\pi$.

\Datafile s may also be plotted as well as functions, in which case the
filename of the \datafile\ to be plotted should be enclosed in either single or
double quotation marks. An example of this syntax would be:

\begin{verbatim}
plot 'data.dat' with points
\end{verbatim}

\noindent which would plot data from the file {\tt data.dat}.
Section~\ref{sec:plot_datafiles} provides further details of the format that
input \datafile s should take and how PyXPlot may be directed to plot only
certain portions of \datafile s.

Multiple datasets can be plotted on a single graph by specifying them in a
comma-separated list, as in the example:

\begin{verbatim}
plot sin(x) with colour blue, cos(x) with linetype 2
\end{verbatim}


\subsection{axes}\indcmd{plot axes}

The {\tt axes} modifier may be used to specify which axes data should be
plotted against when plots have multiple parallel axes -- for example, if a
plot has an {\tt x}-axis along its lower edge and an {\tt x2}-axis along its
upper edge. The following example would plot data using the {\tt x2}-axis as
the ordinate axis and the {\tt y}-axis as the abscissa axis:

\begin{verbatim}
plot sin(x) axes x2y1
\end{verbatim}

\noindent It is also possible to use the {\tt axes} modifier to specify that a
vertical ordinate axis and a horizontal abscissa axis should be used; the
following example would plot data using the {\tt y}-axis as the ordinate axis
and the {\tt x}-axis as the abscissa axis:

\begin{verbatim}
plot sin(x) axes yx
\end{verbatim}


\subsection{title}\indcmd{plot title}

By default, PyXPlot generates its own entry in the legend of a plot for each
dataset plotted.  This default behaviour can be overridden using the {\tt
title} modifier. The following example labels a dataset as `Dataset~1':

\begin{verbatim}
plot sin(x) title 'Dataset 1'
\end{verbatim}

\noindent If a blank string, i.e.\ {\tt ""}, is supplied, then no entry is made
in the plot's legend for that dataset. The same effect can be achieved using
the {\tt notitle} modifier.


\subsection{with}\indcmd{plot with}

The {\tt with} modifier controls the style in which data should be plotted. For
example, the statement
\begin{verbatim}
plot "data.dat" index 1 using 4:5 with lines
\end{verbatim}
specifies that data should be plotted using lines connecting each \datapoint to
its neighbours. More generally, the {\tt with} modifier can be followed by a
range of settings which fine-tune the manner in which the data are displayed;
for example, the statement
\begin{verbatim}
plot "data.dat" with lines linewidth 2.0
\end{verbatim}
would use twice the default width of line.

The following is a complete list of all of PyXPlot's plot styles -- i.e.\ all
of the words which may be used in place of {\tt lines}: {\tt arrows\_\-head},
{\tt arrows\_\-no\-head}, {\tt arrows\_\-two\-head}, {\tt boxes}, {\tt dots},
{\tt Filled\-Region}, {\tt fsteps}, {\tt histeps}, {\tt impulses}, {\tt lines},
{\tt Lines\-Points}, {\tt Lower\-Limits}, {\tt points}, {\tt stars}, {\tt
steps}, {\tt Upper\-Limits}, {\tt wbox\-es}, {\tt X\-Error\-Bars}, {\tt
X\-Error\-Range}, {\tt XY\-Error\-Bars}, {\tt XY\-Error\-Range}, {\tt
XYZ\-Error\-Bars}, {\tt XYZ\-Error\-Range}, {\tt XZ\-Error\-Bars}, {\tt
XZ\-Error\-Range}, {\tt Y\-Error\-Bars}, {\tt Y\-Error\-Range}, {\tt
Y\-Error\-Shaded}, {\tt YZ\-Error\-Bars}, {\tt YZ\-Error\-Range}, {\tt
Z\-Error\-Bars}, {\tt Z\-Error\-Range}. In addition, {\tt lp} and {\tt pl} are
recognised as abbreviations for {\tt lines\-points}; {\tt error\-bars} is
recognised as an abbreviation for {\tt y\-error\-bars}; {\tt error\-range} is
recognised as an abbreviation for {\tt y\-error\-range}; and {\tt
arrows\_\-two\-way} is recognised as an alternative for {\tt
arrows\_\-two\-head}.

As well as the names of these plot styles, the {\tt with} modifier can also be
followed by style modifiers such as {\tt line\-width} which alter the exact
appearance of various plot styles. A complete list of these is as follows:
\begin{itemize}
\item \indmodt{colour} -- used to select the colour in which each dataset is to be plotted. It should be followed either by an integer, to set a colour from the present palette (see Section~\ref{sec:palette}), or by a recognised colour name, a complete list of which can be found in Section~\ref{sec:colour_names}. Alternatively, arbitrary colours may be specified by using one of the forms {\tt rgb0.1:\-0.2:\-0.3}, {\tt hsb0.1:\-0.2:\-0.3} or {\tt cmyk0.4:\-0.3:\-0.2:\-0.1}, where the colon-separated values indicate the RGB, HSB or CMYK components of the desired colour in the range~0 to~1. This modifier may also be spelt {\tt color}.\index{colours!setting for datasets}
\item \indmodt{fillcolour} -- used to select the colour in which each dataset is filled. The colour may be specified using any of the styles listed for {\tt colour}. May also be spelt {\tt fillcolour}.
\item \indmodt{linetype} -- used to numerically select the type of line -- for example, solid, dotted, dashed, etc.\ -- which should be used in line-based plot styles. A complete list of PyXPlot's numbered line types can be found in Chapter~\ref{ch:linetypes_table}. May be abbreviated {\tt lt}.
\item \indmodt{linewidth} -- used to select the width of line, where one represents the default width, which should be used in line-based plot styles. May be abbreviated {\tt lw}.
\item \indmodt{pointlinewidth} -- used to select the width of line, where one represents the default width, which should be used to stroke points in point-based plot styles. May be abbreviated {\tt plw}.
\item \indmodt{pointsize} -- used to select the size of drawn points, where one represents the default size. May be abbreviated {\tt ps}.
\item \indmodt{pointtype} -- used to numerically select the type of point -- for example, crosses, circles, etc.\ -- used by point-based plot styles. A complete list of PyXPlot's numbered point types can be found in Chapter~\ref{ch:linetypes_table}. May be abbreviated {\tt pt}.
\end{itemize}

Any number of these modifiers may be placed sequentially after the {\tt with}
keyword, as in the following examples:

\begin{verbatim}
plot 'datafile' using 1:2 with points pointsize 2
plot 'datafile' using 1:2 with lines colour red linewidth 2
plot 'datafile' using 1:2 with lp col 1 lw 2 ps 3
\end{verbatim}

\noindent Where modifiers take numerical values, expressions of the form {\tt
\$2+1}, similar to those supplied to the {\tt using} modifier, may be used to
indicate that each datapoint should be displayed in a different style or in a
different colour. The following example would plot a \datafile\ with {\tt
points}, drawing the position of each point from the first two columns of the
supplied \datafile\ and the size of each point from the third column:
\begin{verbatim}
plot 'datafile' using 1:2 with points pointsize $3
\end{verbatim}


\section{print}\indcmd{print}

\begin{verbatim}
print <expression> {, <expression>}
\end{verbatim}

The \indcmdt{print} displays a string or the value of a mathematical expression to the
terminal. It is most often used to find the value of a variable, though it can
also be used to produce formatted textual output from a PyXPlot script. For example,

\begin{verbatim}
print a
\end{verbatim}

\noindent would print the value of the variable {\tt a}, and

\begin{verbatim}
print "a = %s"%(a)
\end{verbatim}

\noindent would produce the same result in the midst of formatted text.


\section{pwd}\indcmd{pwd}

\begin{verbatim}
pwd
\end{verbatim}

The \indcmdt{pwd} prints the location of the current working directory.


\section{quit}\indcmd{quit}

\begin{verbatim}
quit
\end{verbatim}

The \indcmdt{quit} can be used to exit PyXPlot. See {\tt exit} for more
details.


\section{rectangle}\indcmd{rectangle}

\begin{verbatim}
rectangle [ item <id> ] at <x>, <y> width <w> height <h>
               [ rotate <r> ] [ with {<option>} ]

rectangle [ item <id> ] from <x1>, <y1> to <x2>, <y2>
               [ rotate <r> ] [ with {<option>} ]
\end{verbatim}

See {\tt box}.


\section{refresh}\indcmd{refresh}

\begin{verbatim}
refresh
\end{verbatim}

The \indcmdt{refresh} produces an exact copy of the latest display. It can be
useful, for example, after changing the terminal type, to produce a second copy
of a plot in a different graphic format. It differs from the {\tt replot}
command in that it doesn't replot anything; use of the {\tt set} command since
the previous {\tt plot} command has no effect on the output.


\section{point}\indcmd{point}

\begin{verbatim}
point [ item <id> ] [at] <x>, <y> [label <string>]
                    [with {<option>} ]
\end{verbatim}

The \indcmdt{point} allows a single point to be plotted on the current
multiplot canvas independently of any graph.  It is equivalent to plotting a
\datafile\ containing a single datum and with invisible axes.  If an optional
{\tt label} is specified then the text string provided is rendered next to the
point.  The {\tt with} modifier allows the style of the point to be specified
using similar options to those accepted by the {\tt plot} command.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{replot}\indcmd{replot}

\begin{verbatim}
replot [ item <id> ] ...
\end{verbatim}

The \indcmdt{replot} may be used to add more datasets to an existing plot, or
to change its axis ranges. For example, having plotted one \datafile\ using the
command

\begin{verbatim}
plot 'datafile1.dat'
\end{verbatim}

\noindent another can be plotted on the same axes using the command

\begin{verbatim}
replot 'datafile2.dat' using 1:3
\end{verbatim}

\noindent or the ranges of the axes on the original plot can be changed using
the command

\begin{verbatim}
replot [0:1][0:1]
\end{verbatim}

\noindent The plot is also updated to reflect any changes to settings made
using the {\tt set} command.  In multiplot mode, the \indcmdt{replot} can
likewise be used to modify the last plot added to the page. For example, the
following would change the title of the latest plot to `foo', and add a plot of
the function $g(x)$ to it:

\begin{verbatim}
set title 'foo'
replot cos(x)
\end{verbatim}

Additionally, it is possible to modify any plot on the page by adding an {\tt
item} modifier to the {\tt replot} statement to specify which plot should be
replotted.  The following example would produce two plots, and then add an
additional function to the first plot:

\begin{verbatim}
set multiplot
plot f(x)
set origin 10,0
plot g(x)
replot item 1 h(x)
\end{verbatim}

If no {\tt item} number is specified, then the \indcmdt{replot} acts by default
upon the most recent plot to have been added to the multiplot canvas.


\section{reset}\indcmd{reset}

\begin{verbatim}
reset
\end{verbatim}

The \indcmdt{reset} returns the values of all settings that have been changed
with the {\tt set} command back to their default values. It also clears the
current multiplot canvas.


\section{save}\indcmd{save}

\begin{verbatim}
save '<filename>'
\end{verbatim}

The \indcmdt{save} saves a list of all of the commands which have been
executed in the current interactive PyXPlot session into a given file. The
filename of the desired location for this file should be placed in quotes, for
example:

\begin{verbatim}
save 'foo'
\end{verbatim}

\noindent would save a command history into the file named {\tt foo}.


\section{set}\indcmd{set}

\begin{verbatim}
set <option> <value>
\end{verbatim}

The \indcmdt{set} sets the value of various operational parameters within
PyXPlot.  For example:

\begin{verbatim}
set pointsize 2
\end{verbatim}

\noindent would sets the default point size to 2. The basic syntax always
follows that above: the {\tt set} command should be followed by some keyword
specifying which setting it is which should be set.  Settings which work in an
on/off fashion tend to take a syntax along the lines of:

\vspace{3mm}
\begin{tabular}{ll}
{\tt set key} & {\tt\it \# Set option ON} \\
{\tt set nokey} & {\tt\it \# Set option OFF}
\end{tabular}
\vspace{3mm}

\noindent
More details of the functions of each individual setting can be found in the
subsections below, which forms a complete list of the recognised setting
keywords.

The reader should also see the {\tt show} command, which can be used to display
the current values of settings, and the {\tt unset} command, which returns
settings to their default values. Chapter~\ref{ch:configuration} describes how
commonly used settings can be saved into a configuration file.


\subsection{arrow}\indcmd{set arrow}

\begin{verbatim}
set arrow <arrow number>
      from [<system>] <x>, [<system>] <y>
      to   [<system>] <x>, [<system>] <y>
      [with {<option>} ]
\end{verbatim}

\noindent where {\tt <system>} is may take any of the values
\newline\noindent
{\tt ( first | second | screen | graph | axis<number> )}
\vspace{5mm}

The \indcmdt{set arrow} causes an arrow to be added to a plot. The example

\begin{verbatim}
set arrow 1 from 0,0 to 1,1
\end{verbatim}

\noindent would draw an arrow between the points $(0,0)$ and $(1,1)$, as
measured on the {\tt x} and {\tt y}-axes.  The tag {\tt 1} immediately following the {\tt
arrow} keyword is an identification number, and allows the arrow to be removed
later with the {\tt unset arrow} command.  By default the coordinates are
measured relative to the first horizontal and vertical axes, but can be specified in a range
of coordinate systems. These are specified as in the example:

\begin{verbatim}
set arrow 1 from first 0, second 0 to axis3 1, axis4 1
\end{verbatim}

\noindent As can be seen, the name of the desired coordinate system precedes the position
value in that coordinate system. The coordinate system {\tt first}, the default,
measures the graph using the {\tt x}- and {\tt y}-axes. {\tt second} uses the {\tt x2}- and
{\tt y2}-axes.  {\tt screen} and {\tt graph} both measure in centimetres from the
origin of the graph.  The syntax {\tt axisn} may also be
used, to use the $n$th horizontal or vertical axis; for example, {\tt axis3} above.

The {\tt set arrow} command can be followed by the keyword {\tt with}, to
specify the style of the arrow. For example, the specifiers {\tt nohead}, {\tt
head} and {\tt twohead}, after the keyword {\tt with}, can be used to make
arrows with no arrow heads, normal arrow heads, or two arrow heads. {\tt twoway}
is an alias for {\tt twohead}.  Normal line type modifiers can also be used
here.  For example:

\begin{verbatim}
set arrow 2 from first 0, second 2.5 to axis3 0,
             axis4 2.5 with colour blue nohead
\end{verbatim}


\subsection{autoscale}\indcmd{set autoscale}

\begin{verbatim}
set autoscale {<axis>}
\end{verbatim}

The {\tt autoscale} setting causes PyXPlot to choose the scaling for an axis
automatically based on the data and/or functions to be plotted against it. The example

\begin{verbatim}
set autoscale x1
\end{verbatim}

\noindent would cause the size of the first horizontal axis to be scaled to fit the
data.  Multiple axes can be specified, viz.:

\begin{verbatim}
set autoscale x1y3
\end{verbatim}

\noindent Note that ranges explicitly specified in a \indcmdt{plot} will override
the autoscale setting.


\subsection{axescolour}\indcmd{set axescolour}

\begin{verbatim}
set axescolour <colour>
\end{verbatim}

The {\tt axescolour} setting changes the colour of the plot's axes.  For example:

\begin{verbatim}
set axescolour blue
\end{verbatim}

\noindent changes the axes to be blue. Any of the recognised colour names listed in
Section~\ref{sec:colour_names} can be used.


\subsection{axis}\indcmd{set axis}

\begin{verbatim}
set axis <axis> [ ( visible | invisible ) ]
  [ ( top | bottom | left | right | front | back ) ]
  [ ( atzero | notatzero ) ]
  [ ( automirrored | mirrored | fullmirrored ) ]
  [ ( noarrow | arrow | reversearrow | twowayarrow ) ]
  [ linked [item <number>] <axis> [using <expression>] ]
\end{verbatim}

The \indcmdt{set axis} may be used to add additional axes to plots and to
configure their appearance. Where an axis is stated on its own, as in the
example
\begin{verbatim}
set axis x2
\end{verbatim}
additional horizontal or vertical axes are added with default
settings. The converse statements
\begin{verbatim}
set noaxis x2
unset axis x2
\end{verbatim}
are used, respectively, to remove axes from plots and to return them
to their default configurations, which often has the same effect of removing
them from the graph, unless they are configured otherwise in a configuration
file.

The position of any axis can be explicitly set using syntax of the form:
\begin{verbatim}
set axis x top
set axis y right
set axis z back
\end{verbatim}
Horizontal axes can be set to appear either at the {\tt top} or {\tt bottom};
vertical axes can be set to appear either at the {\tt left} or {\tt right}; and
$z$-axes can be set to appear either at the {\tt front} or {\tt back}.
By default, the {\tt x1} axis is placed along the bottom of graphs and the {\tt
y1} axis is placed up the left-hand side of graphs. On three-dimensional plots,
the {\tt z1} axis is placed at the front of the graph. The second set of axes
are placed opposite to the first: the {\tt x2}, {\tt y2} and {\tt z2} axes are
placed respectively along the top, right and back sides of graphs.
Higher-numbered axes are placed alongside the {\tt x1}, {\tt y1} and {\tt z1}
axes.

The following keywords may also be placed alongside the positioning keywords
listed above to specify how the axis should appear:
\begin{itemize}
\item {\tt arrow} -- Specifies that an arrowhead should be drawn on the right/top end of the axis. [{\bf Not default}].
\item {\tt atzero} -- Specifies that rather than being placed along an edge of the plot, the axis should mark the lines where the perpendicular axes {\tt x1}, {\tt y1} and/or {\tt z1} are zero. [{\bf Not default}].
\item {\tt automirrored} -- Specifies that an automatic decision should be made between the behaviour of {\tt mirrored} and {\tt nomirrored}. If there are no axes on the opposite side of the graph, a mirror axis is produced. If there are already axes on the opposite side of the graph, no mirror axis is produced. [{\bf Default}].
\item {\tt fullmirrored} -- Similar to {\tt mirrored}. Specifies that this axis should have a corresponding twin placed on the opposite side of the graph with mirroring ticks and labelling. [{\bf Not default}; see {\tt automirrored}].
\item {\tt invisible} -- Specifies that the axis should not be drawn; data can still be plotted against it, but the axis is unseen. See Example~\ref{ex:australia} for a plot where all of the axes are invisible.
\item {\tt linked} -- Specifies that the axis should be linked to another axis; see Section~\ref{sec:linked_axes}.
\item {\tt mirrored} -- Specifies that this axis should have a corresponding twin placed on the opposite side of the graph with mirroring ticks but with no labels on the ticks. [{\bf Not default}; see {\tt automirrored}].
\item {\tt noarrow} -- Specifies that no arrowheads should be drawn on the ends of the axis. [{\bf Default}].
\item {\tt nomirrored} -- Specifies that this axis should not have any corresponding twins. [{\bf Not default}; see {\tt automirrored}].
\item {\tt notatzero} -- Opposite of {\tt atzero}; the axis should be placed along an edge of the plot. [{\bf Default}].
\item {\tt notlinked} -- Specifies that the axis should no longer be linked to any other; see Section~\ref{sec:linked_axes}. [{\bf Default}].
\item {\tt reversearrow} -- Specifies that an arrowhead should be drawn on the left/bottom end of the axis. [{\bf Not default}].
\item {\tt twowayarrow} -- Specifies that arrowheadsshould be drawn on both ends of the axis. [{\bf Not default}].
\item {\tt visible} -- Specifies that the axis should be displayed; opposite of {\tt invisible}. [{\bf Default}].
\end{itemize}


\subsection{axisunitstyle}\indcmd{set axisunitstyle}

\begin{verbatim}
set axisunitstyle ( bracketed | squarebracketed | ratio )
\end{verbatim}

The {\tt axisunitstyle} setting controls the style with which the units of
plotted quantities are indicated on the axes of plots. The {\tt bracketed}
option causes the units to be placed in parentheses following the axis labels,
whilst the {\tt squarebracketed} option using square brackets instead.  The
{\tt ratio} option causes the units to follow the label as a divisor so as to
leave the quantity dimensionless.


\subsection{backup}\indcmd{set backup}

\begin{verbatim}
set backup
\end{verbatim}

The setting {\tt backup} changes PyXPlot's behaviour when it detects that a file
which it is about to write is going to overwrite an existing file. Whereas by
default the existing file would be overwritten by the new one, when the
{\tt backup} setting is turned on, it is renamed, placing a tilde at the end of
its filename. For example, suppose that a plot were to be written with filename
{\tt out.ps}, but such a file already existed.  With the backup setting turned on
the existing file would be renamed {\tt out.ps$\sim$} to save it from being overwritten.

The setting may be turned off via {\tt set nobackup}.


\subsection{bar}\indcmd{set bar}

\begin{verbatim}
set bar ( large | small | <value> )
\end{verbatim}

The {\tt bar} setting changes the size of the bar on the end of the error bars,
relative to the current point size.  For example:

\begin{verbatim}
set bar 2
\end{verbatim}

\noindent sets the bars to be twice the size of the points.  The options {\tt large} and
{\tt small} are equivalent to~1 (the default) and~0 (no bar) respectively.


\subsection{binorigin}\indcmd{set binorigin}

\begin{verbatim}
set binorigin <value>
\end{verbatim}

The {\tt binorigin} setting affects the behaviour of the \indcmdt{histogram} by
adjusting where it places the boundaries between the bins into which it places
data. The \indcmdt{histogram} selects a system of bins which, if extended to
infinity in both directions, would put a bin boundary at the value specified in
the \indcmdt{set binorigin}. Thus, if a value of $0.1$ were specified to the
\indcmdt{set binorigin}, and a bin width of~20 were chosen by the
\indcmdt{histogram}, bin boundaries might lie at~$20.1$, $40.1$, $60.1$, and so
on. The specified value may have any physical units, but must be real and
finite.


\subsection{binwidth}\indcmd{set binwidth}

\begin{verbatim}
set binwidth <value>
\end{verbatim}

The {\tt binwidth} setting changes the width of the bins used by the histogram
command. The specified width may have any physical units, but must be real and
finite.


\subsection{boxfrom}\indcmd{set boxfrom}

\begin{verbatim}
set boxfrom <value>
\end{verbatim}

The {\tt boxfrom} setting alters the vertical line from which bars are drawn
when PyXPlot draws bar charts.  By default, bars all originate from the line
$y=0$, but the example

\begin{verbatim}
set boxfrom 2
\end{verbatim}

\noindent would make the bars originate from the line $y=2$. The specified
vertical ordinate may have any physical units, but must be real and finite.



\subsection{boxwidth}\indcmd{set boxwidth}

\begin{verbatim}
set boxwidth <width>
\end{verbatim}

The {\tt boxwidth} setting alters PyXPlot's behaviour when plotting bar charts.
It sets the default width of the boxes used, in ordinate axis units.  If the
specified width is negative then, as happens by default, the boxes have
automatically selected widths, such that the interfaces between them occur at
the horizontal midpoints between their specified positions.  For example:

\begin{verbatim}
set boxwidth 2
\end{verbatim}

\noindent would set all boxes to be two units wide, and

\begin{verbatim}
set boxwidth -2
\end{verbatim}

\noindent would set all of the bars to have differing widths, centred upon
their specified horizontal positions, such that their interfaces occur at the
horizontal midpoints between them. The specified width may have any physical
units, but must be real and finite.


\subsection{calendar}\indcmd{set calendar}

\begin{verbatim}
set calendar [ ( input | output ) ] <calendar>
\end{verbatim}

The \indcmdt{set calendar} sets the calendar that PyXPlot uses internally to
process dates.  PyXPlot understands two separate calendars: an input calendar
for processing dates that the user reads in from files, and an output calendar
that controls dates printed out or written to plots.  The available calendars
are british, french, gregorian, jewish, muslim, russian, catholic, greek,
hebrew, islamic and julian.

\subsection{clip}\indcmd{set clip}

\begin{verbatim}
set clip
\end{verbatim}

The \indcmdt{set clip} causes PyXPlot to clip points which extend over the edge
of plots. The opposite effect may be achieved using the {\tt set noclip}
command.


\subsection{data style}\indcmd{set data style}

See {\tt set style data}.


\subsection{display}\indcmd{set display}

\begin{verbatim}
set [no]display
\end{verbatim}

By default, whenever an item is added to a multiplot canvas, or an existing
item moved or replotted, the whole multiplot is redrawn to reflect the change.
This can be a time-consuming process when constructing large and complex
multiplot canvases, as fresh output is produced at each step. For this reason,
the {\tt set nodisplay} command is provided, which stops PyXPlot from producing
any graphical output. The {\tt set display} command can subsequently be issued
to return to normal behaviour.  To use these, a script which produces a large
and complex multiplot canvas is typically wrapped as follows:

\begin{verbatim}
set nodisplay
...
set display
refresh
\end{verbatim}


\subsection{filter}\indcmd{set filter}

\begin{verbatim}
set filter '<filename wildcard>' '<filter command>'
\end{verbatim}

The \indcmdt{set filter} allows input filter programs to be specified to allow
PyXPlot to deal with file types that are not in its \datafile\ format.  Firstly
the set of \datafile s that the filter should apply to must be specified; the
standard wildcard characters {\tt *} and {\tt ?} may be used.  Then a filter
program should be specified, along with any necessary options.  This should
take the name of the file to be filtered as the final option on its command
line, immediately following any options specified above.  It should and provide
a suitable PyXPlot data file on standard output for PyXPlot to read.  For
example, to filter all files that end in {\tt .foo} through the {\tt defoo}
filter using the {\tt --text} option one would type:

\begin{verbatim}
set filter "*.foo" "/usr/local/bin/defoo --text"
\end{verbatim}

\subsection{fontsize}\indcmd{set fontsize}

\begin{verbatim}
set fontsize <value>
\end{verbatim}

The {\tt fontsize} setting changes the size of the font used to render all text
labels which appear on a plot, including keys, axis labels, etc. The value
specified should be a multiplicative factor greater than zero; a value of~{\tt
2} would cause text to be rendered at twice its normal size, and a value
of~{\tt 0.5} would cause text to be rendered at half its normal size.  The
default value is one.

As an alternative, font sizes can be specified directly in the \LaTeX\ text of
labels, as in the example:

\begin{verbatim}
set xlabel '\Large This is a BIG label'
\end{verbatim}


\subsection{function style}\indcmd{set function style}

See {\tt set style function}.


\subsection{grid}\indcmd{set grid}

\begin{verbatim}
set [no]grid {<axis>}
\end{verbatim}

The {\tt grid} setting controls whether a grid is placed behind a plot or not.
Issuing the command

\begin{verbatim}
set grid
\end{verbatim}

\noindent would cause a grid to be drawn with its grid lines connecting to the ticks of
the default axes (usually the first horizontal and vertical axes).  Conversely, issuing the command

\begin{verbatim}
set nogrid
\end{verbatim}

\noindent would remove from the plot all grid lines associated with the ticks of any axes.
One or more axes can be specified for the {\tt set grid} command; a grid will
then be drawn to connect with the ticks of these axes as in the example

\begin{verbatim}
set grid x1 y3
\end{verbatim}

\noindent which would cause gridlines to be drawn from ticks of the first horizontal and third
vertical axes.

It is possible, though not always aesthetically pleasing, to draw gridlines
from multiple parallel axes, for example:

\begin{verbatim}
set grid x1x2x3
\end{verbatim}


\subsection{gridmajcolour}\indcmd{set gridmajcolour}

\begin{verbatim}
set gridmajcolour <colour>
\end{verbatim}

The {\tt gridmajcolour} setting changes the colour that is used to plot the
gridlines (see the {\tt set grid} command) which are associated with the major
ticks of axes (i.e.\ major gridlines). For example:

\begin{verbatim}
set gridmajcolour purple
\end{verbatim}

\noindent would cause the major grid lines to be drawn in purple. Any of the recognised
colour names listed in Section~\ref{sec:colour_names} can be used.

See also the {\tt set gridmincolour} command.


\subsection{gridmincolour}\indcmd{set gridmincolour}

\begin{verbatim}
set gridmincolour <colour>
\end{verbatim}

The {\tt gridmincolour} setting changes the colour that is used to plot the
gridlines (see the {\tt set grid} command) which are associated with the minor
ticks of axes (i.e.\ minor gridlines). For example:

\begin{verbatim}
set gridmincolour purple
\end{verbatim}

\noindent would cause the minor grid lines to be drawn in purple. Any of the recognised
colour names listed in Section~\ref{sec:colour_names} can be used.

See also the {\tt set gridmajcolour} command.


\subsection{key}\indcmd{set key}

\begin{verbatim}
set key <position> [<xoffset>, <yoffset>]
\end{verbatim}

The {\tt key} setting determines whether a legend is placed on a plot, and if
so, where it should be located on the plot. Issuing the command:

\begin{verbatim}
set key
\end{verbatim}

\noindent simply causes a legend to be added to the plot in its default position, usually
the plot's upper-right corner. The converse action is achieved by:

\begin{verbatim}
set nokey
\end{verbatim}

\noindent or:

\begin{verbatim}
unset key
\end{verbatim}

\noindent both of which cause a plot to have no legend. A position for the key may also
be specified after the {\tt set key} command, for example:

\begin{verbatim}
set key bottom left
\end{verbatim}

Recognised positions are {\tt top}, {\tt bottom}, {\tt left}, {\tt
right}, {\tt below}, {\tt above}, {\tt outside}, {\tt xcentre} and
{\tt ycentre}. In addition, if none of these quite achieved the desired
result, a positional offset may be specified after one of the position keywords
above.  The first value is assumed to be a horizontal offset, and the second a
vertical offset, both measured in centimetres.  For example:

\begin{verbatim}
set key bottom left 0.0, -0.5
\end{verbatim}

\noindent would display a key below the bottom left corner of the graph.


\subsection{keycolumns}\indcmd{set keycolumns}

\begin{verbatim}
set keycolumns ( <value> | auto )
\end{verbatim}

The {\tt keycolumns} settings sets how many columns the legend of a plot should
be arranged into. By default, all of the entries in the legends of plots are
arranged in a single vertical list. However, for plots with very large number
of datasets, it may be preferably to split this list into several columns. The
{\tt set keycolumns} command can be followed by any positive integer, as in the
example:

\begin{verbatim}
set keycolumns 3
\end{verbatim}


\subsection{label}\indcmd{set label}

\begin{verbatim}
set label <label number> '<text>'
      [<system>] <x>, [<system>] <y>
      [rotate <angle>]
      [with colour <colour>]
\end{verbatim}

\noindent where {\tt <system>} is may take any of the values
\newline\noindent
{\tt ( first | second | screen | graph | axis<number> )}
\vspace{5mm}

The \indcmdt{set label} can be used to place text labels onto a plot.  For
example:

\begin{verbatim}
set label 1 'Hello' 0, 0
\end{verbatim}

\noindent would place the word `Hello' at plot coordinates $(0,0)$, as measured
on the {\tt x}- and {\tt y}-axes.  The tag {\tt 1} immediately following the {\tt
label} keyword is an identification number, and allows the label to be removed
later with the {\tt unset label} command.  By default the position coordinates
of the label are measured relative to the first horizontal and vertical axes, but can be
specified in a range of coordinate systems. These are specified as in the example:

\begin{verbatim}
set label 1 'Hello' first 0, second 0
\end{verbatim}

As can be seen, the name of the desired coordinate system precedes the position
value in that coordinate system. Following \gnuplot's nomenclature, the
coordinate system {\tt first} the default, measures the graph using the {\tt x}- and
{\tt y}-axes. {\tt second} uses the {\tt x2}- and {\tt y2}-axes.  {\tt screen} and {\tt
graph} both measure in centimetres from the origin of the graph.  The syntax
{\tt axisn} may also be used, to use the $n$th horizontal or vertical axis; for example,
{\tt axis3}:

\begin{verbatim}
set label 1 'Hello' axis3 1, axis4 1
\end{verbatim}

A rotation angle may optionally be specified after the keyword {\tt rotate}
to produce text rotated to any arbitrary angle, measured in degrees
counter-clockwise. The following example would produce upward-running text:

\begin{verbatim}
set label 1 'Hello' 1.2, 2.5 rotate 90
\end{verbatim}

By default the labels are black; however, an arbitrary colour may be specified
using the {\tt with colour} modifier.  For example:

\begin{verbatim}
set label 3 'A purple label' 0, 0 with colour purple
\end{verbatim}

\noindent would place a purple label at the origin.


\subsection{linewidth}\indcmd{set linewidth}

\begin{verbatim}
set linewidth <value>
\end{verbatim}

The \indcmdt{set linewidth} sets the default line width, in units of postscript
points ($1$\,pt = $1/72$\,inch), of the lines used to plot datasets onto graphs
with the {\tt lines} plot style.  For example in the following statement:

\begin{verbatim}
set linewidth 3
plot sin(x) with lines
\end{verbatim}

\noindent lines of three times the default thickness are plotted.  The
{\tt set linewidth} setting only affects plot statements where no line width is
manually specified.


\subsection{logscale}\indcmd{set logscale}

\begin{verbatim}
set logscale {<axis>} [<base>]
\end{verbatim}

The {\tt logscale} setting causes an axis to be laid out with logarithmically,
rather than linearly, spaced intervals.  For example, issuing the command:

\begin{verbatim}
set log
\end{verbatim}

\noindent would cause all of the axes of a plot to be scaled logarithmically. Alternatively
only one, or a selection of axes, can be set to scale logarithmically as
follows:

\begin{verbatim}
set log x1 y2
\end{verbatim}

This would cause the first horizontal and second vertical axes to be scaled logarithmically.
Linear scaling can be restored to all axes via:

\begin{verbatim}
set nolog
\end{verbatim}

\noindent or:

\begin{verbatim}
unset log
\end{verbatim}

\noindent and to only one, or a selection of axes, via:

\begin{verbatim}
set nolog x1 y2
\end{verbatim}

\noindent or:

\begin{verbatim}
unset log x1y2
\end{verbatim}

Optionally, a base may be specified at the end of the {\tt set logscale}
command, to produce axes labelled in logarithms of arbitrary bases.  The default
base is 10.


\subsection{multiplot}\indcmd{set multiplot}

\begin{verbatim}
set multiplot
\end{verbatim}

Issuing the command:

\begin{verbatim}
set multiplot
\end{verbatim}

\noindent causes PyXPlot to enter multiplot mode, which allows many graphs to
be plotted together and displayed side-by-side. See Section~\ref{sec:multiplot}
for a full discussion of multiplot mode.


\subsection{mxtics}\indcmd{set mxtics}

See {\tt set xtics}.


\subsection{mytics}\indcmd{set mytics}

See {\tt set xtics}.


\subsection{mztics}\indcmd{set mztics}

See {\tt set ztics}.


\subsection{noarrow}\indcmd{set noarrow}

\begin{verbatim}
set noarrow [<arrow number>]
\end{verbatim}

Issuing the command:

\begin{verbatim}
set noarrow
\end{verbatim}

\noindent removes all arrows produced with the {\tt set arrow} command from the
current plot. Alternatively, individual arrows can be removed using the syntax:

\begin{verbatim}
set noarrow 2
\end{verbatim}

\noindent where the tag {\tt 2} here is the identification number given to
the arrow to be removed when it was initially specified with the {\tt set arrow}
command.


\subsection{noaxis}\indcmd{set noaxis}

\begin{verbatim}
set noaxis [ <axis> {, <axis> } ]
\end{verbatim}

The {\tt set noaxis} command is equivalent to the {\tt unset axis} command. It
should be followed by a comma-separated lists of axes, which are to be removed
from the current axis configuration.


\subsection{nobackup}\indcmd{set nobackup}

See {\tt backup}.


\subsection{noclip}\indcmd{set noclip}

See {\tt clip}.


\subsection{nodisplay}\indcmd{set nodisplay}

See {\tt display}.


\subsection{nogrid}\indcmd{set nogrid}

\begin{verbatim}
set nogrid {<axis>}
\end{verbatim}

Issuing the command {\tt set nogrid} removes gridlines from the current plot. On
its own, the command removes all gridlines from the plot, but alternatively,
those gridlines connected to the ticks of certain axes can selectively be
removed.  The following example would remove gridlines associated with the
first horizontal axis and the second vertical axis:

\begin{verbatim}
set nogrid x1 y2
\end{verbatim}


\subsection{nokey}\indcmd{set nokey}

\begin{verbatim}
set nokey
\end{verbatim}

Issuing the command {\tt set nokey} causes plots to be generated with no legend.
See the {\tt set key} command for more details.


\subsection{nolabel}\indcmd{set nolabel}

\begin{verbatim}
set nolabel {<label number>}
\end{verbatim}

Issuing the command:

\begin{verbatim}
set nolabel
\end{verbatim}

\noindent removes all text labels, as set using the {\tt set label} command,
from the current plot. Alternatively, individual labels can be removed using
the syntax:

\begin{verbatim}
set nolabel 2
\end{verbatim}

\noindent where the tag {\tt 2} here is the identification number given to
the label to be removed when it was initially set using the {\tt set label}
command.


\subsection{nologscale}\indcmd{set nologscale}

\begin{verbatim}
set nologscale {<axis>}
\end{verbatim}

The {\tt logscale} setting causes an axis to be laid out with logarithmically,
rather than linearly, spaced intervals. Conversely, the {\tt nologscale} setting
is used to restore linear scaling. For example, issuing the command:

\begin{verbatim}
set nolog
\end{verbatim}

\noindent would cause all of the axes of a plot to be scaled linearly. Alternatively only one,
or a selection of axes, can be set to scale linearly as follows:

\begin{verbatim}
set nologscale x1 y2
\end{verbatim}

This would cause the first horizontal and second vertical axes to be scaled linearly.


\subsection{nomultiplot}\indcmd{set nomultiplot}

\begin{verbatim}
set nomultiplot
\end{verbatim}

The command {\tt set nomultiplot} causes PyXPlot to leave multiplot mode;
whereas in multiplot mode, galleries of plots and vector graphics can be placed
alongside one another, outside of multiplot, only single graphs and vector
graphics objects are displayed at any one time.  See
Section~\ref{sec:multiplot} for a full discussion of multiplot mode.


\subsection{nostyle}\indcmd{set nostyle}

\begin{verbatim}
set nostyle <style number>
\end{verbatim}

The {\tt nostyle} setting deletes a numbered plot style. For example, the command:

\begin{verbatim}
set nostyle 3
\end{verbatim}

\noindent would delete the third plot style, if defined. See the command {\tt set
style} for more details.


\subsection{notitle}\indcmd{set notitle}

\begin{verbatim}
set notitle
\end{verbatim}

Issuing the command {\tt set notitle} will cause graphs to be produced with no
title at the top.


\subsection{noxtics}\indcmd{set noxtics}

\begin{verbatim}
set no<axis>tics
\end{verbatim}

This command causes graphs to be produced with no major tick marks along the
specified axis. For example, the {\tt set noxtics} command removes all major
tick marks from the {\tt x}-axis.


\subsection{noytics}\indcmd{set noytics}

Similar to the {\tt set noxtics} command, but acts on vertical axes.


\subsection{noztics}\indcmd{set noztics}

Similar to the {\tt set noxtics} command, but acts on $z$-axes.


\subsection{numerics}\indcmd{set numerics}

\begin{verbatim}
set numerics [ ( complex | real ) ] [ errors ( explicit | quiet) ]
    [ display ( latex | natural | typeable) ]
    [ sigfig <precision> ]
\end{verbatim}

The \indcmdt{set numerics} allows the way in which numerical quantities are displayed to be adjusted:

\begin{itemize}

\item The {\tt complex} option causes complex arithmetic to be carried out; the {\tt
real} option causes all numbers to be real and any calculation that generates
complex numbers will generate an error.

\item The {\tt errors} option controls how errors are communicated to the user.  The
{\tt explicit} option causes an error message to be printed to the screen each
time a calculation causes an error.  The {\tt quiet} option causes such
calcluations to silently generate a {\tt nan} (not a number) result.  The {\tt
nan} option is a synonym for {\tt quiet}.

\item The {\tt display} option changes the output of numbers that are printed to the
terminal.  Setting the option to {\tt typeable} causes the numbers to be printed
in a form suitable for pasting back into PyXPlot commands.  The {\tt latex} setting
causes \LaTeX-compatible output to be generated.  The {\tt natural} setting
generates concise, readable output which has neither of the above properties.

\item The {\tt sigfig} setting changes the number of significant figures to which
numbers are displayed when printed out on the PyXPlot terminal.  Regardless of
the value set, all calculations are carried out internally at double precision.

\end{itemize}


\subsection{origin}\indcmd{set origin}

\begin{verbatim}
set origin <x>, <y>
\end{verbatim}

The \indcmdt{set origin} sets the location of the bottom-left corner of the
next graph to be placed on a multiplot canvas.  For example, the command:

\begin{verbatim}
set origin 3,5
\end{verbatim}

\noindent would cause the next graph to be plotted with its bottom-left corner
at position $(3,5)$ centimetres on the multiplot canvas. Alternatively, either
of the coordinates may be specified as quantities with physical units of
length, such as {\tt unit(35*mm)}.  The {\tt set origin} command is of little
use outside of multiplot mode.


\subsection{output}\indcmd{set output}

\begin{verbatim}
set output '<filename>'
\end{verbatim}

The {\tt output} setting controls the name of the file that is produced for
non-interactive terminals ({\tt postscript}, {\tt eps}, {\tt jpeg}, {\tt gif}
and {\tt png}).  For example:

\begin{verbatim}
set output 'myplot.eps'
\end{verbatim}

\noindent causes the output to be written to the file {\tt myplot.eps}.


\subsection{palette}\indcmd{set palette}

\begin{verbatim}
set palette <colour> {, <colour>}
\end{verbatim}

PyXPlot has a palette of colours which it assigns sequentially to datasets when
colours are not manually assigned. This is also the palette to which reference
is made if the user issues a command such as:

\begin{verbatim}
plot sin(x) with colour 5
\end{verbatim}

\noindent requesting the fifth colour from the palette. By default, this palette
contains a range of distinctive colours. However, the user can choose to
substitute his own list of colours using the {\tt set palette} command. It
should be followed by a comma-separated list of colour names, for example:

\begin{verbatim}
set palette red,green,blue
\end{verbatim}

If, after issuing this command, the following plot statement were to be
executed:

\begin{verbatim}
plot sin(x), cos(x), tan(x), exp(x)
\end{verbatim}

\noindent the first function would be plotted in red, the second in green, and the third
in blue. Upon reaching the fourth, the palette would cycle back to red.

Any of the recognised colour names listed in Section~\ref{sec:colour_names} can be used.


\subsection{papersize}\indcmd{set papersize}

\begin{verbatim}
set papersize ( <named size> | <height>,<width> )
\end{verbatim}

The {\tt papersize} option sets the size of output produced by the postscript
terminal. This can take the form of either a recognised paper size name -- a
list of these is given in Appendix~\ref{ch:paper_sizes} -- or as a (height,
width) pair of values, both measured in millimetres. The following examples
demonstrate these possibilities:

\begin{verbatim}
set papersize a4
set papersize letter
set papersize 200,100
\end{verbatim}


\subsection{pointlinewidth}\indcmd{set pointlinewidth}

\begin{verbatim}
set pointlinewidth <value>
\end{verbatim}

The {\tt pointlinewidth} setting changes the width of the lines that are used to
plot \datapoint s.  For example,

\begin{verbatim}
set pointlinewidth 20
\end{verbatim}

\noindent would cause points to be plotted with lines 20 times the default thickness.

The {\tt pointlinewidth} setting can be abbreviated as {\tt plw}.


\subsection{pointsize}\indcmd{set pointsize}

\begin{verbatim}
set pointsize <value>
\end{verbatim}

The {\tt pointsize} setting changes the size at which points are plotted
relative to their default size. It should be followed by a single value, the
relative size, which can be any positive number. For example:

\begin{verbatim}
set pointsize 1.5
\end{verbatim}

\noindent would cause points to be plotted 1.5 times the default size.


\subsection{preamble}\indcmd{set preamble}

\begin{verbatim}
set preamble <text>
\end{verbatim}

The {\tt preamble} setting changes the preamble that is prepended to each item of
text rendered using \LaTeX{}.  This allows, for example, different packages to
be loaded by default and user-defined macros to be set up.


%\subsection{projection}\indcmd{set projection}
%
%\begin{verbatim}
%set projection flat
%\end{verbatim}
%
%The {\tt set projection} command will change between different projections of
%plots in a future version of PyXPlot.  Currently only the {\tt flat} option is
%supported.


\subsection{samples}\indcmd{set samples}

\begin{verbatim}
set samples <value>
\end{verbatim}

The {\tt samples} setting determines the number of values along the ordinate
axis at which functions are evaluated when they are plotted. For example:

\begin{verbatim}
set samples 100
\end{verbatim}

\noindent would cause functions to be evaluated at 100~points along the
ordinate axis.  Increasing this value will cause functions to be plotted more
smoothly, but also more slowly, and the postscript files generated will also be
larger. When functions are plotted with the {\tt points} plot style, this
setting controls the number of points plotted.


\subsection{seed}\indcmd{set seed}

\begin{verbatim}
set seed <value>
\end{verbatim}

The {\tt set seed} command sets the seed used by all of those mathematical
functions which generate random samples from probability distributions.  This
allows repeatable sequences of pseudo-random numbers to be generated.


\subsection{size}\indcmd{set size}

\begin{verbatim}
set size ( <width> | ratio <ratio> | noratio | square )
\end{verbatim}

The setting {\tt size} is used to set the width or aspect ratio of the next
graph to be generated. If a width is specified, then it may either take the
form of a dimensionless number implicitly measured in centimetres, or a
quantity with physical dimensions of length such as {\tt unit(50*mm)}. When the
keyword {\tt ratio} is specified, it should be followed by the ratio of the
graph's height to its width. The keyword {\tt noratio} returns the aspect ratio
to its default value of the golden ratio, and the keyword {\tt square} sets the
aspect ratio to one.


\subsubsection{noratio}\index{set size command!noratio modifier@{\tt noratio} modifier}

\begin{verbatim}
set size noratio
\end{verbatim}

Executing the command:

\begin{verbatim}
set size noratio
\end{verbatim}

\noindent resets PyXPlot to produce plots with its default aspect ratio, which is the
golden section. Other aspect ratios can be set with the {\tt set size ratio}
command.


\subsubsection{ratio}\index{set size command!ratio modifier@{\tt ratio} modifier}

\begin{verbatim}
set size ratio <ratio>
\end{verbatim}

This command sets the aspect ratio of plots produced by PyXPlot.  The height of
resulting plots will equal the plot width, as set by the {\tt set width}
command, multiplied by this aspect ratio.  For example:

\begin{verbatim}
set size ratio 2.0
\end{verbatim}

\noindent would cause PyXPlot to produce plots that are twice as high as they
are wide.  The default aspect ratio which PyXPlot uses is a golden ratio of
$2/(1+\sqrt{5})$.


\subsubsection{square}\index{set size command!square modifier@{\tt square} modifier}

\begin{verbatim}
set size square
\end{verbatim}

This command sets PyXPlot to produce square plots, i.e.\ with unit aspect
ratio. Other aspect ratios can be set with the {\tt set size ratio} command.


\subsection{style}\indcmd{set style}

\begin{verbatim}
set style <style number> {<style option>}
\end{verbatim}

At times, the string of style keywords following the {\tt with} modifier in
plot commands can grow rather unwieldily long. For clarity, frequently used
plot styles can be stored as numbered plot {\tt styles}. The syntax for setting
a plot style is:

\begin{verbatim}
set style 2 points pointtype 3
\end{verbatim}

\noindent where the {\tt 2} is the identification number of the plot style.
In a subsequent plot statement, this line style can be recalled as follows:

\begin{verbatim}
plot sin(x) with style 2
\end{verbatim}


\subsection{style data | style function}\indcmd{set style data}\indcmd{set style function}

\begin{verbatim}
set style { data | function } {<style option>}
\end{verbatim}

The {\tt set style data} command affects the default style with which data from
files is plotted.  Likewise the {\tt set style function} command changes the
default style with which functions are plotted.  Any valid style modifier can
be used.  For example:

\begin{verbatim}
set style data points
set style function lines linestyle 1
\end{verbatim}

\noindent would cause \datafile s to be plotted by default using points and
functions using lines with the first defined line style.


\subsection{terminal}\indcmd{set terminal}

\begin{verbatim}
set terminal ( X11_SingleWindow | X11_MultiWindow | X11_Persist |
               bmp | eps | gif | jpeg | pdf | png | postscript |
               svg | tiff )
             ( colour | color | monochrome )
             ( dpi <value> )
             ( portrait | landscape )
             ( invert | noinvert )
             ( transparent | solid )
             ( antialias | noantialias )
             ( enlarge | noenlarge )
\end{verbatim}

The {\tt set terminal} command controls the graphic format in which PyXPlot
should output plots, for example setting whether it should output plots to files
or display them in a window on the screen. Various options can also be set
within many of the graphic formats which PyXPlot supports using this command.

The following graphic formats are supported:  {\tt X11\_SingleWindow}, {\tt
X11\_MultiWindow}, {\tt X11\_Persist}, {\tt bmp}, {\tt eps}, {\tt gif}, {\tt
jpeg}, {\tt pdf}, {\tt png}, {\tt postscript}, {\tt svg}\footnote{The {\tt svg}
output terminal is experimental and may be unstable. It relies upon the use of
the {\tt svg} output device in Ghostscript, which may not be present on all
systems.}, {\tt tiff}.  To select one of these formats, simply type the name of
the desired format after the {\tt set terminal} command. To obtain more details
on each, see the subtopics below.

The following settings, which can also be typed following the {\tt set
terminal} command, are used to change the options within some of these graphic
formats: {\tt colour}, {\tt monochrome}, {\tt dpi}, {\tt portrait}, {\tt
landscape}, {\tt invert}, {\tt noinvert}, {\tt transparent}, {\tt solid}, {\tt
enlarge}, {\tt noenlarge}. Details of each of these can be found below.


\subsubsection{antialias}\index{set terminal command!antialias modifier@{\tt antialias} modifier}

The {\tt antialias} terminal option causes plots produced with the bitmap
terminals ({\tt bmp}, {\tt gif}, {\tt jpeg}, {\tt png} and {\tt tiff}) to be
antialiased; this is the default behaviour.


\subsubsection{bmp}\index{set terminal command!bmp modifier@{\tt gif} modifier}

The {\tt bmp} terminal renders output as Windows bitmap images. The filename to
which output is to be sent should be set using the {\tt set output} command;
the default is {\tt pyxplot.bmp}. The number of dots per inch used can be
changed using the dpi option. The {\tt invert} option may be used to produce an
image with inverted colours.


\subsubsection{colour}\index{set terminal command!colour modifier@{\tt colour} modifier}

The {\tt colour} terminal option causes plots to be produced in colour; this is
the default behaviour.


\subsubsection{color}\index{set terminal command!color modifier@{\tt color} modifier}

The {\tt color} terminal option is the US-English equivalent of {\tt colour}.


\subsubsection{dpi}\index{set terminal command!dpi modifier@{\tt dpi} modifier}

When PyXPlot is set to produce bitmapped graphics output, using the {\tt bmp},
{\tt gif}, {\tt jpg} or {\tt png} terminals (see the {\tt set terminal}
command), the {\tt dpi} setting changes the number of dots per inch with which
these graphics files are produced. That is to say, it changes the image
resolution of these file formats:

\begin{verbatim}
set terminal dpi 100
\end{verbatim}

\noindent sets the output to a resolution of~100 dots per inch. Higher dpi
values yield better quality images, but larger file sizes.


\subsubsection{enlarge}\index{set terminal command!enlarge modifier@{\tt enlarge} modifier}

The {\tt enlarge} terminal option causes the complete plot to be enlarged or
shrunk to fit the current paper size.


\subsubsection{eps}\index{set terminal command!eps modifier@{\tt eps} modifier}

Sends output to eps files.  The filename to which output is to be sent should
be set using the {\tt set output} command; the default is
{\tt pyxplot.eps}.  This terminal produces encapsulated postscript
suitable for including in, for example, \LaTeX documents.


\subsubsection{gif}\index{set terminal command!gif modifier@{\tt gif} modifier}

The {\tt gif} terminal renders output as gif images. The filename to which
output is to be sent should be set using the {\tt set output} command; the
default is {\tt pyxplot.gif}. The number of dots per inch used can be changed
using the dpi option. Transparent gifs can be produced with the {\tt
transparent} option. The {\tt invert} option may be used to produce an image
with inverted colours.


\subsubsection{invert}\index{set terminal command!invert modifier@{\tt invert} modifier}

The {\tt invert} terminal option causes the bitmap terminals ({\tt bmp}, {\tt
gif}, {\tt jpeg}, {\tt png} and {\tt tiff}) to produce output with inverted
colours. This is useful for producing plots for slideshows, where bright
colours on a dark background may be desired.


\subsubsection{jpeg}\index{set terminal command!jpeg modifier@{\tt jpeg} modifier}

The {\tt jpeg} terminal renders output as jpeg images. The filename to which
output is to be sent should be set using the {\tt set output} command; the
default is {\tt pyxplot.jpg}.  The number of dots per inch used can be changed
using the dpi option. The {\tt invert} option may be used to produce an
image with inverted colours.


\subsubsection{landscape}\index{set terminal command!landscape modifier@{\tt landscape} modifier}

The {\tt landscape} terminal option causes PyXPlot's output to be displayed in
rotated orientation.  This can be useful for fitting graphs onto sheets of
paper, but is generally less useful for plotting things on screen.


\subsubsection{monochrome}\index{set terminal command!monochrome modifier@{\tt monochrome} modifier}

The {\tt monochrome} terminal option causes plots to be rendered in black and
white; by default, different dash styles are used to differentiate between
lines on plots with several datasets.


\subsubsection{noantialias}\index{set terminal command!noantialias modifier@{\tt noantialias} modifier}

The {\tt noantialias} terminal option causes plots produced with the bitmap
terminals ({\tt bmp}, {\tt gif}, {\tt jpeg}, {\tt png} and {\tt tiff}) not to
be antialiased.


\subsubsection{noenlarge}\index{set terminal command!noenlarge modifier@{\tt noenlarge} modifier}

The {\tt noenlarge} terminal option causes the output not to be scaled (the
opposite of {\tt enlarge} above).


\subsubsection{noinvert}\index{set terminal command!noinvert modifier@{\tt noinvert} modifier}

The {\tt noinvert} terminal option causes the bitmap terminals ({\tt gif}, {\tt
jpeg}, {\tt png}) to produce normal output without inverted colours. The
converse of {\tt inverse}.


\subsubsection{pdf}\index{set terminal command!pdf modifier@{\tt pdf}
modifier}

The {\tt pdf} terminal renders output in Adobe's portable document format
(pdf).


\subsubsection{png}\index{set terminal command!png modifier@{\tt png} modifier}

The {\tt png} terminal renders output as png images. The filename to which
output is to be sent should be set using the {\tt set output} command; the
default is {\tt pyxplot.png}. The number of dots per inch used can be changed
using the dpi option. Transparent pngs can be produced with the {\tt
transparent} option. The {\tt invert} option may be used to produce an image
with inverted colours.


\subsubsection{portrait}\index{set terminal command!portrait modifier@{\tt portrait} modifier}

The {\tt portrait} terminal option causes PyXPlot's output to be displayed in
upright (normal) orientation; it is the converse of the {\tt landscape} option.


\subsubsection{postscript}\index{set terminal command!postscript modifier@{\tt postscript} modifier}

Sends output to postscript files. The filename to which output is to be sent
should be set using the {\tt set output} command; the default is {\tt
pyxplot.ps}.  This terminal produces non-encapsulated postscript suitable for
sending directly to a printer.


\subsubsection{solid}\index{set terminal command!solid modifier@{\tt solid} modifier}

The solid option causes the {\tt gif} and {\tt png} terminals to produce output
with a non-transparent background, the converse of {\tt transparent}.


\subsubsection{transparent}\index{set terminal command!transparent modifier@{\tt transparent} modifier}

The {\tt transparent} terminal option causes the {\tt gif} and {\tt png}
terminals to produce output with a transparent background.


\subsubsection{X11\_multiwindow}\index{set terminal command!X11\_multiwindow modifier@{\tt X11\_multiwindow} modifier}

Displays plots on the screen in X11 windows. Each time a new
plot is generated it appears in a new window, and the old plots remain visible.
As many plots as may be desired can be left on the desktop simultaneously.

\subsubsection{X11\_persist}\index{set terminal command!X11\_persist
modifier@{\tt X11\_persist} modifier}

Displays plots on the screen in X11 windows.  Each time a new
plot is generated it appears in a new window, and the old plots remain visible.
When PyXPlot is exited the windows remain in place until they are closed
manually.

\subsubsection{X11\_singlewindow}\index{set terminal command!X11\_singlewindow modifier@{\tt X11\_singlewindow} modifier}

Displays plots on the screen in X11 windows. Each time a new
plot is generated it replaces the old one, preventing the desktop from becoming
flooded with old plots. This terminal is the default when running
interactively.


\subsection{textcolour}\indcmd{set textcolour}

\begin{verbatim}
set textcolour <colour>
\end{verbatim}

The {\tt textcolour} setting changes the colour of all text displayed on a plot.
For example:

\begin{verbatim}
set textcolour red
\end{verbatim}

\noindent causes all text labels, including the labels on graph axes and
legends, etc. to be rendered in red. Any of the recognised colour names listed
in Section~\ref{sec:colour_names} can be used, as can a number which indexes
into the current palette.


\subsection{texthalign}\indcmd{set texthalign}

\begin{verbatim}
set texthalign ( left | centre | right )
\end{verbatim}

The {\tt texthalign} setting controls how text labels are justified
horizontally with respect to their specified positions, acting both upon labels
placed on plots using the {\tt set label} command, and upon text items created
using the {\tt text} command. Three options are available:

\begin{verbatim}
set texthalign left
set texthalign centre
set texthalign right
\end{verbatim}


\subsection{textvalign}\indcmd{set textvalign}

\begin{verbatim}
set textvalign ( bottom | centre | top )
\end{verbatim}

The {\tt textvalign} setting controls how text labels are justified vertically
with respect to their specified positions, acting both upon labels placed on
plots using the {\tt set label} command, and upon text items created using the
{\tt text} command. Three options are available:

\begin{verbatim}
set textvalign bottom
set textvalign centre
set textvalign top
\end{verbatim}


\subsection{title}\indcmd{set title}

\begin{verbatim}
set title '<title>'
\end{verbatim}

The {\tt title} setting can be used to set a title for a plot, to be displayed
above it.  For example, the command:

\begin{verbatim}
set title 'foo'
\end{verbatim}

\noindent would cause a title `foo' to be displayed above a graph. The easiest
way to remove a title, having set one, is via:

\begin{verbatim}
unset title
\end{verbatim}


\subsection{trange}\indcmd{set trange}

\begin{verbatim}
set trange <range>
\end{verbatim}

The {\tt set trange} command changes the range of the free parameter {\tt t}
used when generating parametric plots.  For more details of the syntax of the
range specifier, see the {\tt set xrange} command.

\subsection{unit}\indcmd{set unit}

\begin{verbatim}
set unit [ angle ( dimensionless | nodimensionless ) ]
         [ of <dimension> <unit> ]
         [ scheme <unit scheme> ]
         [ preferred <unit> ]
         [ nopreferred <unit> ]
         [ display ( full | abbreviated | prefix | noprefix ) ]
\end{verbatim}

The \indcmdt{set unit} controls how quantities with physical units are
displayed by PyXPlot. The \indcmdt{set unit scheme} provides the most general
configuration option, allowing one of several {\it units
schemes}\index{units!unit schemes} to be selected, each of which comprises a
list of units which are deemed to be members of the particular scheme. For
example, in the CGS unit scheme\index{CGS units}\index{units!CGS}, all lengths
are displayed in centimetres, all masses are displayed in grammes, all energies
are displayed in ergs, and so forth.  In the imperial unit
scheme\index{imperial units}\index{units!imperial}, quantities are displayed in
British imperial units -- inches, pounds, pints, and so forth -- and in the US
unit scheme, US customary units are used. The available schemes are: {\tt
ancient}, {\tt cgs}, {\tt imperial}, {\tt planck}, {\tt si}, and {\tt us}.

To change the unit used to display a particular dimension of quantity the {\tt
set unit of} form should be used.  For example, the following command will
cause all lengths to be displayed in inches:

\begin{verbatim}
set unit of length inch
\end{verbatim}

The \indcmdt{set unit preferred} offers a slightly more flexible way of
achieving the same result. Whereas the \indcmdt{set unit of} can only operate
on named quanities such as lengths and powers, and cannot act upon compound
units such as {\tt W/Hz}, the \indcmdt{set unit preferred} can act upon any
unit or combination of units:
\begin{verbatim}
set unit preferred parsec
set unit preferred W/Hz
set unit preferred Nm
\end{verbatim}
The latter two examples are particularly useful when working with spectral
densities (powers per unit frequency) or torques (forces multiplied by
distances). Unfortunately, both of these units are dimensionally equal to
powers, and so are displayed by PyXPlot in Joules by default. The above
statement overrides such behaviour. Having set a particular unit to be
preferred, this can be unset as in the following example:
\begin{verbatim}
set unit nopreferred parsec
\end{verbatim}

By default, units are displayed in their abbreviated forms, for example {\tt A}
instead of {\tt amperes} and {\tt W} instead of {\tt watts}. Furthermore, SI
prefixes such as milli- and kilo- are applied to SI units where they are
appropriate. Both of these behaviours can be turned on or off, in the former
case with the commands:

\begin{verbatim}
set unit display abbreviated
set unit display full
\end{verbatim}

\noindent and in the latter case using the following pair of commands:

\begin{verbatim}
set unit display prefix
set unit display noprefix
\end{verbatim}


\subsection{width}\indcmd{set width}

\begin{verbatim}
set width <value>
\end{verbatim}

The {\tt width} setting is used to set the width of the next graph to be
generated. The width may be specified either as a dimensionless number
implicitly measured in centimetres, or as a quantity with physical dimensions
of length such as {\tt unit(50*mm)}.


\subsection{xformat}\indcmd{set xformat}

\begin{verbatim}
set <axis>format ( auto | '<format>' )
      ( horizontal | vertical | rotate <angle> )
\end{verbatim}

By default, major ticks are labelled with representations of the ordinate
values at each point, each accurate to the number of significant figures
specified using the \indcmdt{set numerics sigfig}. These labels may appear as
decimals, such as $3.142$, in scientific notion, as in $3\times10^8$, or, on
logarithmic axes where a base has been specified for the logarithms using
syntax such as \footnote{Note that the {\tt x} axis must be referred to as {\tt
x1} here to distinguish this statement from {\tt set log x2}.}
\begin{verbatim}
set log x1 2
\end{verbatim}
in a format such as $1.5\times2^8$.

The \indcmdt{set xformat} -- together with its companions such as {\tt set
yformat} -- may be used to specify a format for the axis labels to take, as
demonstrated by the following pair of examples:
\begin{verbatim}
set xformat "%.2f"%(x)
set yformat "%s$^\prime$"%(y/unit(feet))
\end{verbatim}
The first example specifies that ordinate values should be displayed to two
decimal places along the {\tt x}-axis; the second specifies that distances should
be displayed in feet along the {\tt y}-axis. Note that the dummy variable used to
represent the ordinate value is {\tt x}, {\tt y} or {\tt z} depending upon the
direction of the axis, but that the dummy variable used in the {\tt set
x2format} command is still {\tt x}. The following pair of examples both have
the equivalent effect of returning the {\tt x2}-axis to its default system of
tick labels:
\begin{verbatim}
set x2format auto
set x2format "%s"%(x)
\end{verbatim}

The following example specifies that ordinate values should be displayed as
multiples of $\pi$:
\begin{verbatim}
set xformat "%s$\pi$"%(x/pi)
plot [-pi:2*pi] sin(x)
\end{verbatim}

Note that where possible, PyXPlot intelligently changes the positions where it
places the ticks along the axis to reflect significant points in the chosen
labelling system.  The extent to which this is possible depends upon the format
string supplied. It is generally easier when continuous-varying numerical
values are substituted into strings, rather than discretely-varying values or
strings.

\subsection{xlabel}\indcmd{set xlabel}

\begin{verbatim}
set <axis>label '<text>'
\end{verbatim}

The {\tt xlabel} setting controls the label placed on the {\tt x}-axis.
For example:

\begin{verbatim}
set xlabel '$x$'
\end{verbatim}

\noindent sets the label on the {\tt x}-axis to `$x$'.  Labels can be placed on
higher axes by inserting their number after the `{\tt x}'; for example:

\begin{verbatim}
set x10label 'foo'
\end{verbatim}

\noindent would label the tenth horizontal axis.

Similarly, labels can be placed on vertical axes as follows:

\begin{verbatim}
set ylabel '$y$'
set y2label 'foo'
\end{verbatim}


\subsection{xrange}\indcmd{set xrange}

\begin{verbatim}
set <axis>range <range> [reverse]
\end{verbatim}

The {\tt xrange} setting controls the range of values along the {\tt x}-axes of
plots.  For function plots, this is also the domain across which the function
will be evaluated.  For example:

\begin{verbatim}
set xrange [0:10]
\end{verbatim}

\noindent sets the first horizontal axis to run from~0 to~10.  Higher numbered
axes may be referred to be inserting their number after the {\tt x}; the ranges
of vertical axes may similarly be set by replacing the {\tt x} with a {\tt y}.
Hence:

\begin{verbatim}
set y23range [-5:5]
\end{verbatim}

\noindent sets the range of the 23rd vertical axis to run from~$-5$ to~5.  To
request a range to be automatically scaled an asterix can be used.  The
following command:

\begin{verbatim}
set xrange [:10][*:*]
\end{verbatim}

\noindent would set the {\tt x}-axis to have an upper limit of 10, but does not
affect the lower limit; its range remains at its previous setting.  The first
y-axis is automatically scaled on both its upper and lower limits.

The keyword {\tt reverse} may be used to indicate that the two limits of an
axis should be swapped. This is useful for setting auto-scaling axes to be
displayed running from right to left, or from top to bottom.


\subsection{xtics}\indcmd{set xtics}

\begin{verbatim}
set [m]<axis>tics
    [ ( axis | border | inward | outward | both ) ]
    [ ( auto
          | [<minimum>,] <increment> [, <maximum>]
          | { '<label>' <position> }
         ] )
\end{verbatim}

By default, PyXPlot places a series of tick marks at significant points along
each axis, with the most significant points being labelled.  Labelled tick
marks are termed {\it major} ticks, and unlabelled tick marks are termed {\it
minor} ticks.  The position and appearance of the major ticks along the {\tt
x}-axis can be configured with the \indcmdt{set xtics}; the corresponding
command {\tt set mxtics} configures the appearance of the minor ticks along the
{\tt x}-axis. Analogous commands such as {\tt set ytics} and {\tt set mx2tics}
configure the major and minor ticks along other axes.

The keywords \indkeyt{inward}, \indkeyt{outward} and \indkeyt{both} are used to
configure how the ticks appear -- whether they point inward, towards the plot,
as is default, or outwards towards the axis labels, or in both directions.  The
keyword \indkeyt{axis} is an alias for \indkeyt{inward}, and \indkeyt{border}
an alias for \indkeyt{outward}.

The remaining options are used to configure where along the axis ticks are
placed. If a series of comma-separated values {\tt <minimum>, <increment>,
<maximum>} are specified, then ticks are placed at evenly spaced intervals
between the specified limits. The {\tt <minimum>} and {\tt <maximum>} values
are optional; if only one value is specified then it is taken to be the step
size between ticks. If two values are specified, then the first is taken to be
{\tt <minimum>}. In the case of logarithmic axes, {\tt <increment>} is applied
as a multiplicative step size.

Alternatively, if a bracketed list of quoted tick labels and tick positions are
provided, then ticks can be placed on an axis individually and each given its
own textual label. The quoted tick labels may be omitted, in which case they
are automatically generated:
\begin{verbatim}
set xtics ("a" 1, "b" 2, "c" 3)
set xtics (1,2,3)
\end{verbatim}
The keyword \indkeyt{autofreq} overrides any manual selection of ticks which
may have been placed on an axis and resumes the automatic placement of ticks
along the axis. The \indcmdt{show xtics}, together with its companions such as
{\tt show x2tics} and {\tt show ytics}, may be used to query the current
ticking options. The \indcmdt{set noxtics} may be used to stipulate that no
ticks should appear along a particular axis:

\begin{verbatim}
set noxtics
show xtics
\end{verbatim}


\subsection{yformat}\indcmd{set yformat}

See {\tt xformat}.


\subsection{ylabel}\indcmd{set ylabel}

See {\tt xlabel}.


\subsection{yrange}\indcmd{set yrange}

See {\tt xrange}.


\subsection{ytics}\indcmd{set ytics}

See {\tt xtics}.


\subsection{zformat}\indcmd{set zformat}

See {\tt xformat}.


\subsection{zlabel}\indcmd{set zlabel}

See {\tt xlabel}.


\subsection{zrange}\indcmd{set zrange}

See {\tt xrange}.


\subsection{ztics}\indcmd{set ztics}

See {\tt xtics}.


\section{show}\indcmd{show}

\begin{verbatim}
show ( all | axes | functions | settings | units
       | userfunctions | variables | <parameter> ... )
\end{verbatim}

The \indcmdt{show} displays the present state of settings which can be set with
the {\tt set} command. For example,

\begin{verbatim}
show pointsize
\end{verbatim}

\noindent will display the current default point size.

Details of the various settings that can be shown can be found under the {\tt
set} command; any keyword which can follow the {\tt set} command can also follow
the {\tt show} command.

In addition, {\tt show all} shows the configuration state of all aspects of
PyXPlot. The command {\tt show settings} shows all of PyXPlot's settings, as
distinct from variables, functions and axes. {\tt show axes} shows the
configuration of all of PyXPlot's axes. {\tt show variables} lists all of the
currently defined variables. And finally, {\tt show functions} lists all of the
current user-defined functions.


\section{solve}\indcmd{solve}

\begin{verbatim}
solve <equation> {, <equation>}
    via <variable> {, <variable>}
\end{verbatim}

The \indcmdt{solve} can be used to solve simple systems of one or more
equations numerically. It takes as its arguments a comma-separated list of the
equations which are to be solved, and a comma-separated list of the variables
which are to be solved for. The latter should be prefixed by the word {\tt
via}, to separate it from the list of equations.

Note that the time taken by the solver dramatically increases with the number
of variables which are simultaneously solved for, meanwhile the accuracy
achieved simultaneously decreases. The following example solves a simple pair
of simultaneous equations of two variables:

\begin{verbatim}
pyxplot> solve x+y=10, x-y=3 via x,y
pyxplot> print x
6.5
pyxplot> print y
3.5
\end{verbatim}

\noindent No output is returned to the terminal if the numerical solver
succeeds, otherwise an error message is displayed. If any of the fitting
variables are already defined prior to the \indcmdt{solve} being called, their
values are used as initial guesses, otherwise an initial guess of unity for
each fitting variable is assumed. Thus, the same \indcmdt{solve} returns two
different values in the following two cases:

\begin{verbatim}
pyxplot> x= # Undefine x
pyxplot> solve cos(x)=0 via x
pyxplot> print x/pi
0.5
pyxplot> x=10
pyxplot> solve cos(x)=0 via x
pyxplot> print x/pi
3.5
\end{verbatim}

\noindent In cases where any of the variables being solved for are not
dimensionless, it is essential that an initial guess with appropriate units be
supplied, otherwise the solver will try and fail to solve the system of
equations using dimensionless values:

\begin{verbatim}
x = unit(m)
y = 5*unit(km)
solve x=y via x
\end{verbatim}

The {\tt solve} command works by minimising the squares of the residuals of all
of the equations supplied, and so even when no exact solution can be found, the
best compromise is returned. The following example has no solution -- a system
of three equations with two variables is over-constrained -- but PyXPlot
nonetheless finds a compromise solution:

\begin{verbatim}
pyxplot> solve x+y=10, x-y=3, 2*x+y=16 via x,y
pyxplot> print x
6.3571429
pyxplot> print y
3.4285714
\end{verbatim}

When complex arithmetic is enabled, the {\tt solve} command allows each of the
variables being fitted to take any value in the complex plane, and thus the
number of dimensions of the fitting problem is effectively doubled -- the real
and imaginary components of each variable are solved for separately -- as in
the following example:

\begin{verbatim}
pyxplot> set numerics complex
pyxplot> solve exp(x)=e*i via x
pyxplot> print Re(x)
1
pyxplot> print Im(x)/pi
0.5
\end{verbatim}


\section{spline}\indcmd{spline}

\begin{verbatim}
spline [ <range> ] <function name>"()" '<filename>'
       [ every <expression> {:<expression>} ]
       [ index <value> ]
       [ select <expression> ]
       [ using <expression>  {:<expression>} ]
\end{verbatim}

The \indcmdt{spline} is an alias for the {\tt interpolate spline} command.
See the entry for the {\tt interpolate} command for more details.


\section{swap}\indcmd{swap}

\begin{verbatim}
swap <item1> <item2>
\end{verbatim}

Items on multiplot canvases are drawn in order of increasing identification
number, and thus items with low identification numbers are drawn first, at the
back of the multiplot, and items with higher identification numbers are later,
towards the front of the multiplot. When new items are added, they are given
higher identification numbers than previous items, and appear at the front of
the multiplot.

If this ordering is not the desired ordering, then the \indcmdt{swap} may
be used to rearrange items. It simply takes two multiplot items, and swaps
their identification numbers and hence positions in the ordered sequence in
which the items on the multiplot are drawn. Thus, if, for example, the corner
of item~3 disappears behind the corner of item~5, when the converse effect is
actually desired, the following command should be issued:
\begin{verbatim}
swap 3 5
\end{verbatim}


\section{tabulate}\indcmd{tabulate}

\begin{verbatim}
tabulate [ <range> ] ( <expression> | <filename> )
       [ every <expression> {:<expression>} ]
       [ index <value> ]
       [ select <expression> ]
       [ sortby <expression> ]
       [ using <expression>  {:<expression>} ]
       [ with <output format> ]
\end{verbatim}

The \indcmdt{tabulate} produces a text file containing the values of a
function at a set of points.  For example, to produce a \datafile\ called {\tt
sine.dat} with the principal values of the sine function:

\begin{verbatim}
set output 'sine.dat'
tabulate [-pi:pi] sin(x)
\end{verbatim}

The {\tt tabulate} command can also be used to select portions of \datafile s.
For example, to select the third, sixth and ninth columns of the \datafile\ {\tt
data.dat}, but only when the arcsine of the value in the fourth column is
positive:

\begin{verbatim}
set output 'filtered.dat'
tabulate 'data.dat' u 3:6:9 select (asin($4)>0)
\end{verbatim}

The format used in each column of the output file is chosen automatically with
integers and small numbers treated intelligently to produce output which
preserves accuracy, but is also easily human-readable. If desired, however, a
format statement may be specified using the {\tt with format} specifier. The
syntax for this is similar to that expected by the Python string substitution
operator ({\tt \%})\index{\% operator@{\tt \%} operator}\index{string
operators!substitution}\footnote{Note that this operator can also be used
within PyXPlot; see Section~\ref{sec:stringsubop} for details.}.  For example,
to tabulate the values of $x^2$ to very many significant figures one could use:

\begin{verbatim}
tabulate x**2 with format "%27.20e"
\end{verbatim}

If there are not enough columns present in the supplied format statement it
will be repeated in a cyclic fashion; e.g. in the example above the single
supplied format is used for both columns.

The {\tt index}, {\tt every}, {\tt using} and {\tt select} modifiers work in the
same way as for the {\tt plot} command.  For example multiple functions may be
tabulated into the same file with the {\tt using} modifier:

\begin{verbatim}
tabulate [0:2*pi] sin(x):cos(x):tan(x) u 1:2:3:4
\end{verbatim}

The {\tt samples} setting can be used to control the number of points that are
inserted into the \datafile.  If the ordinate axis is set to be logarithmic then the
points at which the functions are evaluated are spaced logarithmically.

The data produced by the {\tt tabulate} command can be sorted in order of any
arbitrary metric by supplying an expression after the {\tt sortby} modifier;
where such expressions are supplied, the data is sorted in order from the
smallest value of the expression to the largest.


\section{text}\indcmd{text}

\begin{verbatim}
text [ item <id> ] '<text string>' [ at <x>, <y> ]
     [ rotate <angle> ] [ gap <gap> ]
     [ halign <alignment> ] [ valign <alignment> ]
     [ with colour <colour> ]
\end{verbatim}

The \indcmdt{text} allows strings of text to a added as labels on multiplot
canvases. The example

\begin{verbatim}
text 'Hello World!' at 0,2
\end{verbatim}

\noindent would render the text `Hello World!' at position $(0,2)$,
measured in centimetres. The alignment of the text item with respect to this
position can be set using the {\tt set texthalign} and {\tt set textvalign}
commands, or via the {\tt halign} and {\tt valign} modifiers supplied to the
{\tt text} command itself.

A gap may be specified, which should either have dimensions of length, or be
dimensionless, in which case it is interpreted as being measured in
centimetres. If a gap is specified, then the text string is slightly displaced
from the specified position, in the direction in which it is being aligned.

A rotation angle may optionally be specified after the keyword {\tt rotate}
to produce text rotated to any arbitrary angle, measured in degrees
counter-clockwise. The following example would produce upward-running text:

\begin{verbatim}
text 'Hello' at 1.5, 3.6 rotate 90
\end{verbatim}

By default the text is black; however, an arbitrary colour may be specified
using the {\tt with colour} modifier.  For example:

\begin{verbatim}
text 'A purple label' at 0, 0 with colour purple
\end{verbatim}

\noindent would add a purple label at the origin of the multiplot.

Outside of multiplot mode, the text command can be used to produce images
consisting simply of one single text item. This can be useful for importing
\LaTeX ed equations as gif images into slideshow programs such as Microsoft
Powerpoint which are incapable of producing such neat mathematical notation
by themselves.

All vector graphics objects placed on multiplot canvases receive unique
identification numbers which count sequentially from one, and which may be
listed using the {\tt list} command.  By reference to these numbers, they can
be deleted and subsequently restored with the {\tt delete} and {\tt undelete}
commands respectively.


\section{undelete}\indcmd{undelete}

\begin{verbatim}
undelete <item number> { , <item number> }
\end{verbatim}

The \indcmdt{undelete} allows vector graphics objects which have previously
been deleted from the current multiplot canvas to be restored.  The item(s)
which are to be restored should be identified using the reference number which
was used to delete it, and as can be found using the {\tt list} command. The
example

\begin{verbatim}
undelete 1
\end{verbatim}

\noindent would cause the previously deleted item numbered {\tt 1} to reappear.


\section{unset}\indcmd{unset}

\begin{verbatim}
unset <setting>
\end{verbatim}

The \indcmdt{unset} causes a setting that has been changed using the set
command to be returned to its default value.  For example:

\begin{verbatim}
unset linewidth
\end{verbatim}

\noindent returns the linewidth to its default value.

The list of keywords which can follow the {\tt unset} command are essentially
the same as those which can follow the {\tt set} command.


\section{while}\indcmd{while}

\begin{verbatim}
while <condition> [loopname <loopname>]
 "{"
    ...
 "}"
\end{verbatim}

The \indcmdt{while} is a loop construct; it causes a set of commands to be
executed repeatedly whilst the specified {\tt condition} remains true. For
example, the following code prints out the low-valued Fibonacci numbers:

\begin{verbatim}
i = 1
j = 1
while (j < 50)
 {
  print j
  i = i + j
  print i
  j = j + i
 }
\end{verbatim}

